---
title: "Adding Plot Features to Existing Plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding Plot Features to Existing Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  echo = TRUE
)
```

## Overview

This vignette demonstrates how to add **subplot features** (census information, team members, plot characteristics, etc.) to existing plots in the database using the user-friendly `add_plot_features()` function.

**What are subplot features?**

Subplot features are attributes that describe plots or census events, stored in the `data_liste_sub_plots` table:
- **People**: Team leader, principal investigator, data manager, etc.
- **Dates**: Census dates (year, month, day)
- **Plot characteristics**: Various plot-specific measurements and metadata

## Prerequisites

```{r}
library(CafriplotsR)
library(dplyr)

# Connect to database (requires write permissions)
con <- call.mydb()
```

## Quick Start Example

```{r}
# Prepare your data
plot_features <- data.frame(
  plot_name = c("Plot-A", "Plot-B", "Plot-C"),
  team_leader = c("John Doe", "Jane Smith", "Bob Wilson"),
  principal_investigator = c("Dr. Smith", "Dr. Smith", "Dr. Jones"),
  census_year = c(2020, 2020, 2021)
)

# Preview what would be added (DRY RUN - always do this first!)
result <- add_plot_features(
  data = plot_features,
  dry_run = TRUE
)

# If the preview looks good, actually add the features
result <- add_plot_features(
  data = plot_features,
  dry_run = FALSE
)
```

## Step-by-Step Workflow

### Step 1: Discover Available Features

First, see what subplot features are available in the database:

```{r}
# Get all available subplot feature types
available_features <- subplot_list()

# View the list
View(available_features)

# Common feature types:
# - team_leader
# - principal_investigator
# - data_manager
# - data_provider
# - census_date (with year, month, day components)
# - additional_people
# - plot_area
# - vegetation_type
```

### Step 2: Prepare Your Data

Your data should have:
1. A **plot identifier column** (`plot_name` or `id_liste_plots`)
2. One or more **feature columns**

```{r}
# Example 1: Simple features
my_data <- data.frame(
  plot_name = c("DJA_PLOT_001", "LOPE_PLOT_023", "IVINDO_PLOT_012"),
  team_leader = c("Dr. Marie Blanc", "John Doe", "Dr. Jean Nguema"),
  principal_investigator = c("Prof. Gilles Dauby", "Dr. Sarah White", "Prof. Pierre Dupont"),
  census_year = c(2022, 2023, 2023),
  census_month = c(11, 5, 9)
)

# Example 2: Multiple people (comma-separated)
my_data <- data.frame(
  plot_name = c("Plot-A", "Plot-B"),
  team_leader = c("John Doe, Jane Smith", "Bob Wilson"),
  data_manager = c("Alice Brown", "Tom Jones, Emma Davis")
)

# Example 3: Using plot IDs instead of names
my_data <- data.frame(
  id_liste_plots = c(1, 2, 3),
  team_leader = c("John Doe", "Jane Smith", "Bob Wilson")
)
```

### Step 3: Add Features (with Dry Run First!)

**Always use `dry_run = TRUE` first** to preview changes:

```{r}
# Dry run - preview what would be added
preview <- add_plot_features(
  data = my_data,
  dry_run = TRUE
)

# Review the preview
print(preview)
```

**Expected output:**
```
═══ Adding Plot Features ═══
ℹ Mode: DRY RUN (preview only)
ℹ Rows to process: 3

─── Step 1: Identifying plot ID column ───
✓ Using 'plot_name' as plot identifier (name)

─── Step 2: Mapping columns to subplot features ───
✓ Mapped 3 feature column(s):
ℹ   team_leader → team_leader
ℹ   principal_investigator → principal_investigator
ℹ   census_year → census_date

─── Step 3: Validating data ───
✓ Validation passed!

─── Step 4: Preparing features for import ───
✓ Prepared 3 features for import

─── Step 5: Preview - Would Add Features ───
ℹ Would add 3 record(s) for feature: team_leader
ℹ Would add 3 record(s) for feature: principal_investigator
ℹ Would add 3 record(s) for feature: census_date

✓ Dry run completed - no changes made
ℹ Run with dry_run = FALSE to actually import
```

### Step 4: Actually Add Features

If the dry run looks good, proceed with the actual import:

```{r}
# Actual import
result <- add_plot_features(
  data = my_data,
  dry_run = FALSE
)

# Check result
if (result$success) {
  cat("✓ Successfully added features!\n")
  cat("  - Records added:", result$n_rows, "\n")
  cat("  - Plots affected:", result$n_plots, "\n")
  cat("  - Feature types:", paste(result$feature_types, collapse = ", "), "\n")
}
```

### Step 5: Verify Features Were Added

Query the plots to confirm the features were added:

```{r}
# Query plots with features
plots_with_features <- query_plots(
  plot_name = c("Plot-A", "Plot-B", "Plot-C"),
  exact_match = TRUE,
  show_multiple_census = TRUE,
  con = con
)

# Check census features
View(plots_with_features$census_features)

# Or query subplot features directly
subplot_features <- query_subplot_features(
  plot_ids = plots_with_features$id_liste_plots,
  format = "wide",
  con = con
)

View(subplot_features)
```

---

## Advanced Usage

### Handling Custom Column Names

The function automatically maps column names, but you can also provide custom mappings:

```{r}
# Your data has non-standard column names
my_data <- data.frame(
  PlotName = c("Plot-A", "Plot-B"),
  TeamLead = c("John Doe", "Jane Smith"),
  PI = c("Dr. Smith", "Dr. Jones"),
  Year = c(2020, 2021)
)

# Interactive mapping (default - will prompt you)
result <- add_plot_features(
  data = my_data,
  interactive = TRUE,
  dry_run = TRUE
)

# Non-interactive with pre-defined mapping
mapping <- list(
  PlotName = "plot_name",
  TeamLead = "team_leader",
  PI = "principal_investigator",
  Year = "census_year"
)

result <- add_plot_features(
  data = my_data,
  column_mapping = mapping,
  interactive = FALSE,
  dry_run = FALSE
)
```

### Adding Multiple People

For people-related features, you can specify multiple people using comma-separated names:

```{r}
plot_features <- data.frame(
  plot_name = c("Plot-A", "Plot-B"),
  team_leader = c("John Doe, Jane Smith", "Bob Wilson"),
  additional_people = c("Alice Brown, Tom Jones, Emma Davis", "Chris Lee")
)

result <- add_plot_features(
  data = plot_features,
  dry_run = FALSE
)

# The function will:
# 1. Split comma-separated names
# 2. Match each name to table_colnam (or prompt to add if not found)
# 3. Create separate records for each person
```

### Adding Census Date Information

Census dates can be specified in multiple ways:

```{r}
# Option 1: Separate year/month/day columns
census_data <- data.frame(
  plot_name = c("Plot-A", "Plot-B"),
  census_year = c(2020, 2021),
  census_month = c(6, 8),
  census_day = c(15, 22)
)

# Option 2: Just year
census_data <- data.frame(
  plot_name = c("Plot-A", "Plot-B"),
  census_year = c(2020, 2021)
)

# The function will handle partial dates (month/day will be NA if not provided)
result <- add_plot_features(
  data = census_data,
  dry_run = FALSE
)
```

### Non-Interactive Mode (for Scripts)

For automated scripts, disable interactive prompts:

```{r}
# Define mapping explicitly
mapping <- list(
  plot_id = "plot_name",
  leader = "team_leader",
  pi = "principal_investigator"
)

result <- add_plot_features(
  data = my_data,
  column_mapping = mapping,
  interactive = FALSE,
  ask_before_update = FALSE,
  verbose = FALSE,
  dry_run = FALSE
)
```

---

## Column Mapping Details

The function uses intelligent column mapping:

### 1. Exact Match
```
Column "team_leader" → feature "team_leader"
```

### 2. Synonym Match
```
Column "PI" → feature "principal_investigator"
Column "TeamLead" → feature "team_leader"
Column "leader" → feature "team_leader"
```

Common synonyms recognized:
- `PI`, `princ_invest`, `investigator` → `principal_investigator`
- `teamlead`, `team_lead`, `leader` → `team_leader`
- `datamanager`, `data_mgr`, `manager` → `data_manager`
- `area`, `surface`, `plot_size` → `plot_area`

### 3. Fuzzy String Match
```
Column "TeamLeader" → feature "team_leader" (similarity: 85%)
Column "Investigator" → feature "principal_investigator" (similarity: 70%)
```

### 4. Interactive Selection

If auto-mapping fails, you'll be prompted:

```
⚠ Could not auto-map column: 'MyCustomColumn'

Select the subplot feature type for column 'MyCustomColumn':
  0. Skip this column
  1. team_leader
  2. principal_investigator
  3. data_manager
  4. data_provider
  ...
  99. Show all features

Your choice:
```

---

## Common Scenarios

### Scenario 1: Adding Team Information After Plot Creation

You've imported plots but forgot to add team members:

```{r}
# Query existing plots
existing_plots <- query_plots(
  locality_name = "Dja Reserve",
  extract_individuals = FALSE,
  con = con
)

# Prepare team info
team_info <- data.frame(
  plot_name = existing_plots$plot_name,
  team_leader = c("John Doe", "Jane Smith", "Bob Wilson"),
  principal_investigator = rep("Dr. Paul Nguema", nrow(existing_plots)),
  data_manager = c("Alice Brown", "Alice Brown", "Tom Jones")
)

# Add team info
result <- add_plot_features(
  data = team_info,
  dry_run = FALSE
)
```

### Scenario 2: Adding Census Date Information

You want to record when each plot was censused:

```{r}
census_dates <- data.frame(
  plot_name = c("Plot-A", "Plot-B", "Plot-C"),
  census_year = c(2020, 2020, 2021),
  census_month = c(6, 8, 3),
  census_day = c(15, 22, 10)
)

result <- add_plot_features(
  data = census_dates,
  dry_run = FALSE
)
```

### Scenario 3: Bulk Update from Excel File

You have an Excel file with plot features to add:

```{r}
# Load data from Excel
library(readxl)
plot_features <- read_excel("plot_features_to_add.xlsx")

# Preview the data
View(plot_features)

# Dry run to check mapping
preview <- add_plot_features(
  data = plot_features,
  interactive = TRUE,
  dry_run = TRUE
)

# If mapping looks good, proceed
result <- add_plot_features(
  data = plot_features,
  interactive = FALSE,  # Use the same mapping as dry run
  dry_run = FALSE
)
```

### Scenario 4: Adding Features with Plot IDs

If you already have plot IDs (e.g., from a query):

```{r}
# Get plot IDs from query
plots <- query_plots(method = "1ha-IRD", con = con)

# Prepare features using IDs
features <- data.frame(
  id_liste_plots = plots$id_liste_plots[1:5],
  team_leader = c("Person A", "Person B", "Person C", "Person D", "Person E")
)

# Add features
result <- add_plot_features(
  data = features,
  plot_id_column = "id_liste_plots",
  dry_run = FALSE
)
```

---

## Troubleshooting

### Issue: "Plots not found in database"

**Solution**: Ensure plot names exactly match database values. Use `exact_match = TRUE` when querying:

```{r}
# Check if plots exist
existing_plots <- query_plots(
  plot_name = c("Plot-A", "Plot-B"),
  exact_match = TRUE,
  con = con
)

if (nrow(existing_plots) == 0) {
  cat("Plots not found! Check spelling.\n")
}
```

### Issue: "Could not auto-map column"

**Solution**: Use interactive mode or provide explicit mapping:

```{r}
# Option 1: Interactive
result <- add_plot_features(data = my_data, interactive = TRUE, dry_run = TRUE)

# Option 2: Explicit mapping
mapping <- list(MyColumn = "team_leader")
result <- add_plot_features(data = my_data, column_mapping = mapping, dry_run = TRUE)
```

### Issue: "Person not found in table_colnam"

**Solution**: The function will prompt you to add new people interactively. Alternatively, add people first:

```{r}
# Add new person to table_colnam
# (You'll be prompted during import if using interactive mode)

# Or check existing people
people <- DBI::dbGetQuery(con, "SELECT * FROM table_colnam")
View(people)
```

### Issue: "Invalid subplot feature type"

**Solution**: Check available features:

```{r}
# See all valid feature types
available <- subplot_list(con = con)
View(available)

# Use exact names from the 'type' column
```

---

## Best Practices

1. **Always dry run first**: Use `dry_run = TRUE` to preview changes before committing

2. **Verify plot names**: Ensure plot names in your data exactly match database values

3. **Use interactive mode initially**: Let the function help you map columns interactively

4. **Save mappings for reuse**: If you have a standard data format, save the column mapping:
   ```r
   # Save mapping for reuse
   my_standard_mapping <- list(
     PlotName = "plot_name",
     TeamLead = "team_leader",
     PI = "principal_investigator"
   )

   # Use in subsequent imports
   result <- add_plot_features(data, column_mapping = my_standard_mapping,
                                interactive = FALSE, dry_run = FALSE)
   ```

5. **Verify after import**: Always query back the data to confirm features were added correctly

6. **Handle people carefully**: For people-related features, ensure names are formatted consistently
   (e.g., "First Last" or "Last, First")

---

## Related Functions

- `subplot_list()` - List all available subplot feature types
- `query_subplot_features()` - Query existing subplot features
- `add_subplot_features()` - Low-level function (used internally)
- `query_plots()` - Query plot metadata and features

---

## Summary

The `add_plot_features()` function provides a user-friendly interface for adding subplot features to existing plots:

1. ✅ **Intelligent column mapping** (exact, synonym, fuzzy, interactive)
2. ✅ **Automatic validation** (plots exist, feature types valid)
3. ✅ **Dry run mode** (preview before committing)
4. ✅ **People linking** (automatic matching/adding to table_colnam)
5. ✅ **Clear feedback** (detailed progress messages)
6. ✅ **Safe operations** (uses existing add_subplot_features internally)

Always start with a dry run, verify the mapping, then proceed with the actual import!
