---
title: "The CafriplotsR Apps at a Glance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The CafriplotsR Apps at a Glance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  echo = TRUE
)
```

## Introduction

CafriplotsR ships ten interactive applications. They cover the whole life cycle
of an inventory dataset — exploring it, standardizing its taxonomy, importing
it, correcting it, and linking it to herbarium specimens — without writing any
R code beyond the one line that launches the app.

Each app opens on the same login screen. What differs is **who may use it**:
three apps can be opened without any credentials, the other seven need your own
account because they write to the database. One of the three, the taxonomic
matching app, goes further still and also runs offline.

## Which app do I need?

| App | Launch with | Access |
|---|---|---|
| Taxonomic name standardization | `launch_taxonomic_match_app()` | public, account, or offline |
| Taxonomic backbone | `launch_taxo_backbone_app()` | public to browse, account to edit |
| Plot querying | `launch_query_plots_app()` | public or account |
| Plot data import | `launch_import_wizard()` | account |
| Plot features and censuses | `launch_feature_wizard()` | account |
| Record-by-record corrections | `launch_data_update_app()` | account |
| Taxa-level trait import | `launch_taxa_traits_import()` | account |
| Herbarium specimen import | `launch_specimen_import_wizard()` | account |
| Specimen identifications | `launch_specimen_identification_app()` | account |
| Individual ↔ specimen linking | `launch_individual_specimen_linking_app()` | account |

## Working without an account

The login screen of the three read-only apps offers a **Connect as public user**
button. It signs you in through a shared, read-only account that reaches the
taxonomy and the species-level traits. Nothing you do in that mode can change
the database, and the editing controls are hidden rather than disabled.

This is enough to:

- standardize your own species list against the Central African backbone,
- browse the backbone, its synonymy and the traits attached to a taxon,
- explore the inventories that their owners have opened to everyone.

It is **not** enough to import data, correct records, or manage specimens. Those
apps do not show the public button at all, because a read-only account cannot
complete a single one of their workflows.

## Working without a network

One app goes further and runs with no database at all: the taxonomic name
standardization app offers a **Use offline (cached backbone)** button. It works
from a copy of the taxonomic backbone saved on your own machine, so you can
standardize a species list from a field station, a plane or any other place
without a route to the server.

Two things are worth knowing before you rely on it:

- The button only appears once a cache exists. The cache is written the first
  time you run the matching app **online**, so run it connected once before you
  leave.
- Offline, you get automatic matching, fuzzy suggestions and manual review.
  The traits enrichment tab is hidden, because traits live in the database and
  nothing local can stand in for them.

**The other nine apps never show this button.** Offline mode leaves the app with
no connection, and none of them has anything to work on in that state — they
import, correct or query live data. If you were expecting the button somewhere
else, that is why it is not there.

## Explore and standardize

### Taxonomic name standardization

```{r}
launch_taxonomic_match_app()
```

Matches your own list of species names against the Central African plant
taxonomic backbone: automatic matching with fuzzy search, manual review of
whatever the matcher could not resolve on its own, and export of the
standardized list. The result gives each of your names a stable taxon
identifier (`idtax_n`), which is what lets you join traits and inventories
later on.

Start here if you arrive with a species list from your own fieldwork. This is
also the one app that runs offline, from a cached copy of the backbone.

### Taxonomic backbone

```{r}
launch_taxo_backbone_app()
```

Browses and manages the taxonomic backbone itself: searching taxa, inspecting
synonymy relationships, and generating the R code for the query you just built
by hand. With an account, it also adds new taxa, updates existing records and
maintains synonymy. As a public user you get the browsing and code-generation
half; the editing controls do not appear.

### Plot querying

```{r}
launch_query_plots_app()
```

An interactive front end to `query_plots()`: filter plots by country, method or
other criteria, see them on a map, drill down to the individual trees, and
export the result. Which plots you see depends on the account you used —
row-level security policies decide, and a public user sees only the inventories
their owners have opened to everyone.

## Import and update

These apps all require your own account, and they only ever touch the plots
your account is entitled to.

### Plot data import

```{r}
launch_import_wizard()
```

The complete import workflow for plots and individual measurements: upload a
file, map its columns to the database fields, validate, preview what will be
written, and execute. It wraps the package's import functions, so the checks
you would otherwise run by hand are applied for you.

### Plot features and censuses

```{r}
launch_feature_wizard()
```

Adds features to plots that already exist — either a new census, with its dates
and the people involved, or arbitrary plot-level features.

### Record-by-record corrections

```{r}
launch_data_update_app()
```

Corrects plot metadata and individual records one at a time. It is the friendly
counterpart to `update_records()`, which is more powerful but expects you to
already know which table a value lives in.

### Taxa-level trait import

```{r}
launch_taxa_traits_import()
```

Imports trait measurements attached to taxa rather than to individual stems:
upload, map your columns onto the trait list, preview, execute.

## Herbarium specimens

### Specimen import

```{r}
launch_specimen_import_wizard()
```

Imports new herbarium specimens from Excel or CSV files.

### Specimen identifications

```{r}
launch_specimen_identification_app()
```

Updates the identifications recorded for specimens, wrapping
`update_ident_specimens()`.

### Individual ↔ specimen linking

```{r}
launch_individual_specimen_linking_app()
```

Creates the links between individual trees and herbarium specimens, using the
herbarium information carried in your individuals dataset.

## Getting an account

Public access covers taxonomy and traits. Working with inventory data — your
own or a colleague's — needs an account, which also determines which plots you
may query and update. Access is granted per user, so get in touch with the
maintainers to have one created.
