---
title: "Above-Ground Biomass Estimation with BIOMASS Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Above-Ground Biomass Estimation with BIOMASS Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  echo = TRUE,
  fig.width = 7,
  fig.height = 5
)
```

## Overview

This vignette demonstrates how to use `plotsdatabase` with the **BIOMASS** package to estimate above-ground biomass (AGB) for permanent forest plots. The `query_plots()` function with `output_style = "permanent_plot"` provides data structures optimized for allometric analyses.

**Key advantages of the integrated workflow:**

1. **Automatic trait enrichment** - Wood density and other traits are pre-integrated
2. **Ready-made height-diameter table** - `$height_diameter` table provided by default
3. **Structured output** - Separate tables for metadata, individuals, and measurements
4. **Taxonomic hierarchies** - Trait inheritance from species → genus → family

## Prerequisites

```{r setup}
library(plotsdatabase)
library(BIOMASS)
library(dplyr)
library(ggplot2)

# Connect to database
con <- call.mydb()
con_taxa <- call.mydb.taxa()
```

---

## Step 1: Extract Plot Data

Use `output_style = "permanent_plot"` to get structured output optimized for biomass analysis:

```{r extract-data}
# Query permanent plots with individual data
plot_data <- query_plots(
  locality_name = c("Lopé", "Ivindo"),  # Example localities
  extract_individuals = TRUE,
  output_style = "permanent_plot",
  con = con
)

# Check structure
names(plot_data)
# $metadata - Plot-level information
# $individuals - Individual tree data with traits
# $height_diameter - Height-diameter pairs (ready for modeling)
# $censuses - Census information
```

The `permanent_plot` output style provides:

- **`$metadata`**: Plot coordinates, area, census dates, investigators
- **`$individuals`**: Complete tree inventory with taxonomic traits
- **`$height_diameter`**: Pre-filtered height-diameter pairs for allometric modeling
- **`$censuses`**: Census metadata with team leaders and dates

---

## Step 2: Explore the Data

### Plot Metadata

```{r explore-metadata}
# View plot information
plot_data$metadata %>%
  select(plot_name, locality, latitude, longitude,
         plot_area_ha, census_date) %>%
  head()
```

### Individual Tree Data

The individuals table includes **automatically enriched traits**:

```{r explore-individuals}
# Check what data we have
str(plot_data$individuals)

# Key columns for biomass:
# - dbh (stem_diameter): Diameter at breast height (cm)
# - species (tax_sp_level): Species name
# - wood_density_mean (taxa_mean_wood_density): Mean wood density (g/cm³)
# - wood_density_sd (taxa_sd_wood_density): SD for error propagation

# Summary of wood density availability
plot_data$individuals %>%
  summarise(
    n_total = n(),
    n_with_wd = sum(!is.na(wood_density_mean)),
    pct_with_wd = round(100 * n_with_wd / n_total, 1),
    mean_wd = mean(wood_density_mean, na.rm = TRUE),
    sd_wd = sd(wood_density_mean, na.rm = TRUE)
  )
```

**Note**: Wood density traits are automatically:
- Retrieved from the taxa database
- Inherited from species → genus → family when species-level data unavailable
- Averaged at plot level when no taxonomic match found (if `traits_to_genera = TRUE`)

### Height-Diameter Data

The `$height_diameter` table is **pre-filtered and ready** for modeling:

```{r explore-hd}
# Height-diameter pairs (automatically filtered)
head(plot_data$height_diameter)

# This table includes:
# - id_n: Individual ID
# - plot_name: Plot identifier
# - tag: Tree tag
# - D: Diameter (dbh in cm)
# - H: Height (tree height in m)
# - POM: Point of measurement (if available)

# Summary
plot_data$height_diameter %>%
  group_by(plot_name) %>%
  summarise(
    n_pairs = n(),
    mean_dbh = round(mean(D), 1),
    mean_height = round(mean(H), 1),
    dbh_range = paste(round(min(D), 1), "-", round(max(D), 1)),
    height_range = paste(round(min(H), 1), "-", round(max(H), 1))
  )
```

---

## Step 3: Build Height-Diameter Models

Use the pre-filtered `$height_diameter` table directly with BIOMASS:

### Option A: Single Model (All Plots)

```{r hd-model-global}
# Fit global height-diameter model
hd_model_global <- BIOMASS::modelHD(
  D = plot_data$height_diameter$D,
  H = plot_data$height_diameter$H,
  method = "weibull",
  useWeight = TRUE
)

# Check model fit
summary(hd_model_global)

# Residual Standard Error for uncertainty propagation
RSE_global <- hd_model_global$RSE
print(paste("RSE:", round(RSE_global, 3)))
```

### Option B: Plot-Specific Models

For plots with sufficient height measurements (>15 pairs recommended):

```{r hd-model-by-plot}
# Build separate models per plot
hd_models_by_plot <- plot_data$height_diameter %>%
  group_by(plot_name) %>%
  filter(n() >= 15) %>%  # Require at least 15 H-D pairs
  group_split() %>%
  lapply(function(plot_df) {
    tryCatch({
      model <- BIOMASS::modelHD(
        D = plot_df$D,
        H = plot_df$H,
        method = "weibull",
        useWeight = TRUE
      )
      list(
        plot_name = unique(plot_df$plot_name),
        model = model,
        RSE = model$RSE,
        n = nrow(plot_df)
      )
    }, error = function(e) {
      NULL  # Return NULL if model fails
    })
  }) %>%
  Filter(Negate(is.null), .)  # Remove failed models

# Summary
cat(sprintf("Built %d plot-specific models\n", length(hd_models_by_plot)))
```

### Visualize H-D Relationships

```{r plot-hd-models, fig.width=8, fig.height=6}
# Plot height-diameter data with fitted curve
ggplot(plot_data$height_diameter, aes(x = D, y = H)) +
  geom_point(aes(color = plot_name), alpha = 0.6, size = 2) +
  geom_smooth(method = "nls",
              formula = y ~ a * (1 - exp(-b * x^c)),
              method.args = list(start = list(a = 40, b = 0.05, c = 1)),
              se = TRUE, color = "black", linewidth = 1.5) +
  labs(
    title = "Height-Diameter Relationship",
    x = "Diameter at breast height (cm)",
    y = "Tree height (m)",
    color = "Plot"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## Step 4: Predict Missing Heights

Most trees don't have height measurements. Use the H-D model to predict them:

```{r predict-heights}
# Prepare data for AGB estimation
agb_data <- plot_data$individuals %>%
  filter(!is.na(dbh)) %>%  # Must have diameter
  mutate(
    # Predict height using global model (or plot-specific if available)
    H_predicted = BIOMASS::retrieveH(
      D = dbh,
      model = hd_model_global
    )$H,
    # Use measured height if available, otherwise predicted
    H_final = ifelse(!is.na(tree_height), tree_height, H_predicted),
    # Assign RSE for uncertainty
    H_RSE = RSE_global
  )

# Check prediction success
agb_data %>%
  summarise(
    n_total = n(),
    n_measured = sum(!is.na(tree_height)),
    n_predicted = sum(is.na(tree_height)),
    pct_predicted = round(100 * n_predicted / n_total, 1)
  )
```

**Using plot-specific models** (if available):

```{r predict-heights-plot-specific, eval=FALSE}
# Create lookup for plot-specific RSE
plot_rse <- data.frame(
  plot_name = sapply(hd_models_by_plot, function(x) x$plot_name),
  RSE_plot = sapply(hd_models_by_plot, function(x) x$RSE)
)

# Predict with plot-specific models where available
agb_data <- plot_data$individuals %>%
  filter(!is.na(dbh)) %>%
  left_join(plot_rse, by = "plot_name") %>%
  mutate(
    # Use plot-specific RSE if available, otherwise global
    H_RSE = ifelse(!is.na(RSE_plot), RSE_plot, RSE_global),
    # Predict heights (would need to loop through plot-specific models)
    H_final = ifelse(!is.na(tree_height), tree_height, H_predicted)
  )
```

---

## Step 5: Estimate Above-Ground Biomass

Now we have all required inputs for BIOMASS:

- **D**: Diameter (dbh)
- **WD**: Wood density (automatically from database!)
- **H**: Height (measured or predicted)
- **Errors**: SD for wood density, RSE for height

### Check Data Completeness

```{r check-completeness}
# Summary of required variables
summary(agb_data[, c("dbh", "wood_density_mean", "H_final", "H_RSE")])

# Check for missing data
agb_data %>%
  summarise(
    missing_dbh = sum(is.na(dbh)),
    missing_wd = sum(is.na(wood_density_mean)),
    missing_height = sum(is.na(H_final)),
    complete_cases = sum(!is.na(dbh) & !is.na(wood_density_mean) & !is.na(H_final))
  )
```

### Single AGB Estimate (No Error Propagation)

Quick estimate without uncertainty:

```{r agb-simple}
# Simple AGB calculation using Chave et al. 2014 equation
agb_data <- agb_data %>%
  filter(!is.na(dbh) & !is.na(wood_density_mean) & !is.na(H_final)) %>%
  mutate(
    AGB_kg = BIOMASS::computeAGB(
      D = dbh,
      WD = wood_density_mean,
      H = H_final
    )
  )

# Plot-level AGB
agb_by_plot <- agb_data %>%
  group_by(plot_name) %>%
  summarise(
    n_trees = n(),
    total_AGB_Mg = sum(AGB_kg) / 1000,  # Convert kg to Mg
    AGB_per_ha = total_AGB_Mg / first(plot_area_ha)  # Standardize by area
  )

print(agb_by_plot)
```

### Monte Carlo Simulation with Error Propagation

**Recommended approach** - propagates uncertainties from all sources:

```{r agb-monte-carlo}
# Run Monte Carlo simulation (1000 iterations) per plot
agb_mc_results <- agb_data %>%
  group_by(plot_name) %>%
  group_split() %>%
  lapply(function(plot_df) {

    # Run Monte Carlo for this plot
    mc_result <- BIOMASS::AGBmonteCarlo(
      D = plot_df$dbh,
      WD = plot_df$wood_density_mean,
      errWD = plot_df$wood_density_sd,  # Wood density uncertainty
      H = plot_df$H_final,
      errH = plot_df$H_RSE,  # Height prediction uncertainty
      Dpropag = "chave2004",  # Diameter measurement error model
      n = 1000  # Number of Monte Carlo iterations
    )

    # Get plot area for standardization
    plot_area <- unique(plot_df$plot_area_ha)

    # Extract results
    list(
      plot_name = unique(plot_df$plot_name),
      n_trees = nrow(plot_df),
      # AGB statistics (Mg/ha)
      AGB_mean = sum(mc_result$AGB_simu[, 1]) / 1000 / plot_area,
      AGB_sd = sd(colSums(mc_result$AGB_simu)) / 1000 / plot_area,
      AGB_q025 = quantile(colSums(mc_result$AGB_simu), 0.025) / 1000 / plot_area,
      AGB_q975 = quantile(colSums(mc_result$AGB_simu), 0.975) / 1000 / plot_area,
      # Individual tree contributions
      tree_contributions = mc_result$AGB_simu[, 1] / 1000
    )
  })

# Convert to data frame
agb_summary <- data.frame(
  plot_name = sapply(agb_mc_results, function(x) x$plot_name),
  n_trees = sapply(agb_mc_results, function(x) x$n_trees),
  AGB_Mg_ha = sapply(agb_mc_results, function(x) x$AGB_mean),
  AGB_sd = sapply(agb_mc_results, function(x) x$AGB_sd),
  AGB_q025 = sapply(agb_mc_results, function(x) x$AGB_q025),
  AGB_q975 = sapply(agb_mc_results, function(x) x$AGB_q975)
)

print(agb_summary)
```

### Visualize Results

```{r plot-agb-results, fig.width=8, fig.height=6}
# AGB estimates with 95% confidence intervals
ggplot(agb_summary, aes(x = reorder(plot_name, AGB_Mg_ha), y = AGB_Mg_ha)) +
  geom_point(size = 4, color = "darkgreen") +
  geom_errorbar(aes(ymin = AGB_q025, ymax = AGB_q975),
                width = 0.3, linewidth = 1, color = "darkgreen") +
  labs(
    title = "Above-Ground Biomass by Plot",
    subtitle = "Error bars show 95% confidence intervals from Monte Carlo simulation",
    x = "Plot",
    y = "AGB (Mg/ha)"
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold")
  )
```

---

## Step 6: Uncertainty Breakdown

Understand sources of uncertainty in AGB estimates:

```{r uncertainty-breakdown}
# Compare uncertainty sources (illustrative example)
# This requires running AGBmonteCarlo multiple times with different error sources

# Example: Compare with and without height uncertainty
agb_no_h_error <- agb_data %>%
  slice(1:10) %>%  # Small subset for illustration
  summarise(
    mean_AGB = mean(BIOMASS::AGBmonteCarlo(
      D = dbh,
      WD = wood_density_mean,
      errWD = wood_density_sd,
      H = H_final,
      errH = 0,  # No height error
      Dpropag = "chave2004",
      n = 100
    )$AGB)
  )

cat("Uncertainty sources in AGB estimation:\n")
cat("1. Diameter measurement error (~5%)\n")
cat("2. Wood density variability (SD in database)\n")
cat("3. Height prediction error (RSE from H-D model)\n")
cat("4. Allometric equation error (Chave et al. 2014)\n")
```

---

## Advanced: Plot-Level Analysis

### Combine with Plot Characteristics

```{r plot-characteristics}
# Join AGB with plot metadata
plot_analysis <- plot_data$metadata %>%
  left_join(agb_summary, by = "plot_name") %>%
  select(
    plot_name, locality, latitude, longitude,
    plot_area_ha, n_trees, AGB_Mg_ha, AGB_sd
  )

# Explore relationships
ggplot(plot_analysis, aes(x = n_trees, y = AGB_Mg_ha)) +
  geom_point(aes(color = locality), size = 4) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(
    title = "Tree Density vs. Above-Ground Biomass",
    x = "Number of trees",
    y = "AGB (Mg/ha)",
    color = "Locality"
  ) +
  theme_minimal()
```

### Size Class Distribution

```{r size-class-agb}
# AGB contribution by size class
size_class_agb <- agb_data %>%
  mutate(
    size_class = cut(dbh,
                     breaks = c(0, 10, 20, 30, 40, 50, 100, Inf),
                     labels = c("0-10", "10-20", "20-30", "30-40",
                               "40-50", "50-100", ">100"))
  ) %>%
  group_by(plot_name, size_class) %>%
  summarise(
    n_trees = n(),
    total_AGB = sum(AGB_kg) / 1000,
    .groups = "drop"
  )

# Stacked bar plot
ggplot(size_class_agb, aes(x = plot_name, y = total_AGB, fill = size_class)) +
  geom_bar(stat = "identity") +
  labs(
    title = "AGB Distribution by Size Class",
    x = "Plot",
    y = "AGB (Mg)",
    fill = "DBH class (cm)"
  ) +
  scale_fill_viridis_d() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## Summary

This vignette demonstrated the complete workflow for AGB estimation using `plotsdatabase` and `BIOMASS`:

1. ✅ **Extract structured data** with `output_style = "permanent_plot"`
2. ✅ **Use pre-filtered H-D data** from `$height_diameter` table
3. ✅ **Build allometric models** with BIOMASS::modelHD()
4. ✅ **Leverage integrated traits** - wood density automatically included
5. ✅ **Estimate AGB with uncertainty** using Monte Carlo simulation
6. ✅ **Visualize and analyze** results at plot and individual level

**Key advantages:**

- **No manual trait matching** - wood density pre-integrated from taxa database
- **Ready-made H-D table** - filtered and formatted for BIOMASS
- **Hierarchical trait inheritance** - genus/family fallback when species unavailable
- **Error propagation** - uncertainties from all sources properly accounted for

---

## Additional Resources

- **BIOMASS package**: Réjou-Méchain et al. (2017) <https://CRAN.R-project.org/package=BIOMASS>
- **Allometric equations**: Chave et al. (2014) <https://doi.org/10.1111/gcb.12629>
- **plotsdatabase queries**: See `vignette("using-query-plots")`
- **Taxonomic matching**: See `vignette("taxonomic-app")`

---

## References

Chave, J., et al. (2014). Improved allometric models to estimate the aboveground biomass of tropical trees. *Global Change Biology*, 20(10), 3177-3190.

Réjou-Méchain, M., et al. (2017). biomass: an R package for estimating above-ground biomass and its uncertainty in tropical forests. *Methods in Ecology and Evolution*, 8(9), 1163-1167.
