---
title: "Database Connections Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Database Connections Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(CafriplotsR)
```

## Introduction

This guide explains how to connect to the CafriplotsR databases, manage credentials securely, and troubleshoot common connection issues.

CafriplotsR uses **two separate PostgreSQL databases**:
1. **Main database** (`plots_transects`): Contains plot, subplot, and individual tree data
2. **Taxa database** (`rainbio`): Contains taxonomic information and species-level traits

## Quick Start

### First-time Connection

```{r quick-start}
library(CafriplotsR)

# Connect to main database (will prompt for credentials)
con_main <- call.mydb()

# Connect to taxa database (uses same credentials)
con_taxa <- call.mydb.taxa()

# Test your connections
print_connection_status()
```

On first use, you'll be prompted to enter:

- **Username**: Your database username
- **Password**: Your database password

Credentials are cached in memory for the duration of your R session.

## Connection Functions

### Main Database: `call.mydb()`

Use this for all plot, subplot, and individual tree data operations.

```{r main-db}
# Basic connection
con <- call.mydb()

# Force new credentials (if you need to switch users)
con <- call.mydb(reset = TRUE)

# Provide credentials directly (not recommended for security)
con <- call.mydb(user = "myuser", pass = "mypassword")
```

### Taxa Database: `call.mydb.taxa()`

Use this for taxonomic queries and species traits.

```{r taxa-db}
# Connect to taxa database
con_taxa <- call.mydb.taxa()
```

**Important**: The taxa database is **read-only** for most users. Write operations are restricted to administrators.


## Managing Credentials

### Option 1: Interactive Prompts (Default)

The simplest approach - enter credentials when prompted:

```{r interactive}
con <- call.mydb()
# Enter username: [your_username]
# Enter password: [your_password]
```

**Pros**: Secure, no stored passwords
**Cons**: Must enter credentials each new R session

### Option 2: Environment Variables (Persistent)

Store credentials in your `.Renviron` file for automatic loading:

```{r setup-creds}
# Run once to set up
setup_db_credentials()
# Follow the prompts to enter username and password
```

Then use with `use_env_credentials = TRUE`:

```{r use-env}
con <- call.mydb(use_env_credentials = TRUE)
```

**Warning**: Credentials are stored in **plain text** in `~/.Renviron`. Only use this on your personal, secure computer.

To remove stored credentials:

```{r remove-creds}
remove_db_credentials()
```

### Option 3: Direct Parameters (Not Recommended)

You can pass credentials directly, but this is **not recommended** as passwords may be visible in your code or history:

```{r direct-creds}
# Avoid this in shared code!
con <- call.mydb(user = "myuser", pass = "mypassword")
```

## Cleaning Up Connections

### Why Clean Connections?

Database connections are limited resources. Failing to close them can cause:

- **Connection exhaustion**: Database refuses new connections
- **Memory leaks**: Unused connections consume resources
- **Stale connections**: Old connections may timeout and cause errors

### When to Clean Up

Try to clean up connections when:

1. You're done working with the database
2. Before closing R/RStudio
3. When switching between users

### How to Clean Up

```{r cleanup}
# Close all connections and clear cached credentials
cleanup_connections()
```

This function:

- Closes the main database connection
- Closes the taxa database connection
- Clears cached credentials from memory


## Checking Connection Status

### Quick Status Check

```{r status}
# See current connection status
print_connection_status()
```

Example output:
```
-- Database Connections Status --
v Main DB: Connected to plots_transects as myuser
v Taxa DB: Connected to rainbio as myuser
```

### Full Diagnostic

For troubleshooting, run a complete diagnostic:

```{r diagnostic}
db_diagnostic()
```

This shows:

- Connection status for both databases
- Configuration details (host, port, database names)
- Connectivity test results
- PostgreSQL version information

## Checking Data Access (Row-Level Security)

The database uses **row-level security (RLS)** to control which plots each user can access.

### View Your Accessible Plots

```{r accessible-plots}
con <- call.mydb()

# See which plots you can access
result <- get_user_accessible_plots(con, "your_username")
print(result)

# Get just the plot IDs as a vector
plot_ids <- result$plot_ids[[1]]
print(plot_ids)
```

### View Your Policies

```{r list-policies}
# See all policies for a user
list_user_policies(con, user = "your_username")

# See all policies on a table
list_user_policies(con, user = NULL, table = "data_liste_plots")
```

### Understanding Access Levels

Policies can grant different operations:

- **SELECT**: Read-only access
- **INSERT**: Can add new records
- **UPDATE**: Can modify existing records
- **DELETE**: Can remove records
- **ALL**: Full access (SELECT, INSERT, UPDATE, DELETE)

## Troubleshooting

### Common Issues and Solutions

#### "Connection refused" or "Could not connect"

**Causes**:

- Network issues
- Database server is down
- Firewall blocking connection

**Solutions**:

1. Check your internet connection
2. Try again in a few minutes
3. Contact database administrator

#### "Authentication failed"

**Causes**:

- Wrong username or password
- Account doesn't exist

**Solutions**:

```{r auth-fix}
# Reset credentials and try again
con <- call.mydb(reset = TRUE)
```

#### "SSL SYSCALL error: EOF detected"

**Causes**:

- Connection was closed but still being used
- Typically happens after closing a Shiny app

**Solutions**:

```{r ssl-fix}
# Clean up and reconnect
cleanup_connections()
con <- call.mydb()
```

#### "Too many connections"

**Causes**:

- Multiple unclosed connections
- Other users consuming connections

**Solutions**:

```{r conn-fix}
# Clean up your connections
cleanup_connections()

# Wait and try again
Sys.sleep(5)
con <- call.mydb()
```

#### Query timeout or "Lost connection"

**Causes**:

- Very large query
- Network instability
- Server overload

**Solutions**:

The package includes automatic retry for transient failures:

```{r retry}
# func_try_fetch automatically retries failed queries
result <- func_try_fetch(con, "SELECT * FROM large_table")
```

#### "Permission denied" or empty results

**Causes**:

- Row-level security restricting access
- User doesn't have policy for requested plots

**Solutions**:

1. Check your accessible plots:

```{r check-access}
get_user_accessible_plots(con, "your_username")
```

2. Contact administrator to request access to additional plots

### Reset Everything

If you're having persistent issues, do a complete reset:

```{r full-reset}
# 1. Clean up all connections
cleanup_connections()

# 2. Restart R session
.rs.restartR()

# 3. Reconnect fresh
library(CafriplotsR)
con <- call.mydb(reset = TRUE)
```

## Best Practices Summary

### Do's

- **Always call `cleanup_connections()`** before closing R
- **Use `print_connection_status()`** to verify connections
- **Check your plot access** with `get_user_accessible_plots()` if queries return empty
- **Use connection pools** in Shiny apps
- **Run `db_diagnostic()`** when troubleshooting
- **Store credentials in `.Renviron`** only on personal secure computers

### Don'ts

- **Don't hardcode passwords** in scripts
- **Don't share credentials** with others
- **Don't leave connections open** indefinitely
- **Don't ignore SSL errors** - clean up and reconnect
- **Don't create multiple connections** when one will do

### Recommended Workflow

```{r workflow}
library(CafriplotsR)

# 1. Connect
con <- call.mydb()

# 2. Verify connection
print_connection_status()

# 3. Check your access if needed
get_user_accessible_plots(con, "your_username")

# 4. Do your work
# ... queries, updates, etc. ...

# 5. Clean up when done
cleanup_connections()
```

## Getting Help

If you encounter issues not covered here:

1. Run `db_diagnostic()` and note the output
2. Check your accessible plots with `get_user_accessible_plots()`
3. Contact the database administrator with:
   - The diagnostic output
   - The exact error message
   - What you were trying to do
