---
title: "Growth and Mortality Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Growth and Mortality Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(CafriplotsR)
```

## Introduction

This vignette demonstrates how to compute tree growth rates and mortality/recruitment statistics for permanent forest plots with multiple censuses.

Two main functions are available:

- `compute_growth()`: Calculate diameter growth rates between censuses
- `compute_mortality()`: Calculate mortality and recruitment rates

Both functions automatically fetch data from the database and only require plot identifiers as input.

## Computing Growth Rates

### Basic Usage

```{r growth-basic}
# Connect to database
con <- call.mydb()

# Compute growth by plot name
growth <- compute_growth(plot_names = "mbalmayo001", con = con)

# Or by plot ID
growth <- compute_growth(plot_ids = c(1, 2, 3), con = con)
```

### Understanding the Output

The function returns a list with two components:

```{r growth-output}
# Summary statistics per plot and census interval
growth$summary

# Individual-level growth data
growth$individuals
```

**Summary columns:**

- `plot_name`: Plot identifier
- `census_pair`: Census interval (e.g., "1-2" for first to second census)
- `n_individuals`: Total individuals measured in both censuses
- `n_valid`: Individuals with valid growth measurements
- `n_excluded`: Individuals excluded due to measurement errors
- `mean_growth_mm_yr`: Mean diameter growth rate (mm/year)
- `sd_growth_mm_yr`: Standard deviation of growth
- `median_growth_mm_yr`: Median growth rate
- `mean_dbh_1_mm`: Mean DBH at first census (mm)
- `mean_dbh_2_mm`: Mean DBH at second census (mm)
- `mean_interval_years`: Mean time between censuses (years)

### Growth Calculation Methods

Two methods are available:

```{r growth-methods}
# Incremental growth (default): (DBH2 - DBH1) / time
growth_inc <- compute_growth(plot_ids = 1, method = "I")

# Exponential growth: (log(DBH2) - log(DBH1)) / time
growth_exp <- compute_growth(plot_ids = 1, method = "E")
```

### Adjusting Quality Control Parameters

```{r growth-params}
growth <- compute_growth(
  plot_ids = 1,
  mindbh = 100,      # Minimum DBH in mm (default: 100)
  err.limit = 4,     # Error limit for negative growth detection (default: 4)
  maxgrow = 75       # Maximum valid growth in mm/year (default: 75)
)
```

**Quality control:**

- Trees with DBH < `mindbh` are excluded
- Extreme negative growth (measurement errors) is detected using a statistical model
- Growth rates > `maxgrow` mm/year are flagged as errors

### Summary Only

To get only summary statistics without individual data:

```{r growth-summary-only}
growth <- compute_growth(
  plot_ids = 1,
  return_individual = FALSE
)
```

## Computing Mortality and Recruitment

### Basic Usage

```{r mortality-basic}
# Compute mortality by plot name
mort <- compute_mortality(plot_names = "mbalmayo001", con = con)

# Or by plot ID
mort <- compute_mortality(plot_ids = c(1, 2, 3), con = con)
```

### Understanding the Output

The function returns a list with three components:

```{r mortality-output}
# Summary statistics
mort$summary

# List of individuals that died
mort$dead_individuals

# List of newly recruited individuals
mort$recruits
```

**Summary columns:**

- `plot_name`: Plot identifier
- `census_pair`: Census interval
- `N_outset`: Number of individuals at start of interval
- `N_dead`: Number of deaths
- `N_survivor`: Number of survivors
- `N_recruits`: Number of new recruits
- `mortality_rate`: Exponential mortality rate
- `mortality_percent_yr`: Annual mortality percentage
- `recruitment_rate`: Exponential recruitment rate
- `recruitment_percent_yr`: Annual recruitment percentage
- `mean_interval_years`: Time between censuses
- `mean_dbh_dead_cm`: Mean DBH of dead trees (cm)
- `mean_dbh_recruit_cm`: Mean DBH of recruits (cm)

### Mortality Rate Calculation

Mortality rate is calculated using the exponential model:

$$\lambda = \frac{\ln(N_0) - \ln(N_s)}{t}$$

Where:
- $N_0$ = number of individuals at start
- $N_s$ = number of survivors
- $t$ = time interval in years

Annual mortality percentage is: $(1 - e^{-\lambda}) \times 100$

### Examining Dead Trees

```{r mortality-dead}
# View dead individuals with their characteristics
mort$dead_individuals

# Filter by taxonomy
library(dplyr)
mort$dead_individuals %>%
  filter(tax_fam == "Fabaceae")
```

### Examining Recruits

```{r mortality-recruits}
# View recruited individuals
mort$recruits

# Size distribution of recruits
hist(mort$recruits$dbh_at_recruitment_cm)
```

## Working with Multiple Plots

Both functions can process multiple plots at once:
```{r multiple-plots}
# Growth for multiple plots
growth <- compute_growth(
  plot_names = c("plot001", "plot002", "plot003")
)

# Results are combined with plot_name column for identification
growth$summary %>%
  group_by(plot_name) %>%
  summarise(
    mean_growth = mean(mean_growth_mm_yr, na.rm = TRUE),
    n_intervals = n()
  )
```

## Complete Workflow Example

```{r workflow}
library(CafriplotsR)
library(dplyr)

# Connect
con <- call.mydb()

# Define plots to analyze
my_plots <- c("site_A_plot1", "site_A_plot2", "site_B_plot1")

# Compute growth
growth <- compute_growth(plot_names = my_plots, con = con)

# View growth summary
growth$summary %>%
  arrange(plot_name, census_pair)

# Compute mortality
mort <- compute_mortality(plot_names = my_plots, con = con)

# Compare mortality across plots
mort$summary %>%
  select(plot_name, census_pair, mortality_percent_yr, recruitment_percent_yr) %>%
  arrange(plot_name)

# Identify species with high mortality
mort$dead_individuals %>%
  count(tax_sp_level, sort = TRUE)

# Clean up
cleanup_connections()
```

## Troubleshooting

### "Need at least 2 censuses"

This error occurs when the selected plots don't have multiple census records:

```{r trouble-census}
# Check if your plot has multiple censuses
query_result <- query_plots(
  plot_names = "your_plot",
  show_multiple_census = TRUE,
  extract_individuals = FALSE
)

# Look at census_features to see available censuses
query_result$census_features
```

### No valid measurements

If growth calculations return few valid measurements, check:

1. **Minimum DBH threshold**: Lower `mindbh` if needed
2. **Date information**: Ensure census dates are recorded in the database
3. **Measurement quality**: Review excluded individuals in the output

```{r trouble-valid}
# Check individual results for exclusions
growth <- compute_growth(plot_ids = 1)
table(growth$individuals$accepted_growth)
```

## References

Growth trimming methodology adapted from:

- CTFS R Package: http://ctfs.si.edu/Public/CTFSRPackage/

Mortality rate calculations follow standard forest dynamics methodology as described in:

- Sheil, D., & May, R. M. (1996). Mortality and recruitment rate evaluations in heterogeneous tropical forests. Journal of Ecology, 84(1), 91-100.
