---
title: "Importing Plot Data into the Database"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Importing Plot Data into the Database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  echo = TRUE
)
```

## Overview

This vignette demonstrates the complete workflow for importing new plot data into the database. The process involves two main steps:

1. **Import plot metadata** - Add plot information (location, size, census dates, etc.)
2. **Import individual tree data** - Add tree measurements and traits for individuals within those plots

Both workflows use **transaction-based imports** with automatic rollback on errors, ensuring data integrity.

## Prerequisites

Before importing data, ensure you have:

1. **Database credentials** configured with write permissions
2. **Taxonomic names standardized** using `launch_taxonomic_match_app()` (see taxonomic-app vignette)
3. **Plot coordinates and metadata** ready
4. **Individual tree data** in Excel or CSV format

```{r}
library(CafriplotsR)
library(dplyr)

# Connect to database
con <- call.mydb()
```

---

## Step 1: Import Plot Metadata

Before importing individual trees, you must first create the plots in the database.

### 1.1 Prepare Plot Data

Plot data should include essential metadata:

```{r}
# Example: Prepare plot metadata
plot_metadata <- data.frame(
  plot_name = c("Plot-A", "Plot-B", "Plot-C"),
  country = c("Gabon", "Gabon", "Cameroon"),
  locality_name = c("Lopé NP", "Lopé NP", "Dja Reserve"),
  ddlat = c(-0.5, -0.52, 3.2),
  ddlon = c(11.5, 11.48, 13.5),
  plot_area = c(1, 1, 0.5),  # hectares
  date_census_1 = as.Date(c("2020-01-15", "2020-02-10", "2019-11-20")),
  team_leader_1 = c("John Doe", "Jane Smith", "Bob Wilson"),
  principal_investigator_1 = c("Dr. Smith", "Dr. Smith", "Dr. Jones"),
  method = c("1ha-IRD", "1ha-IRD", "0.5ha-custom"),
  subplot_shape = c("square", "square", "rectangle"),
  stringsAsFactors = FALSE
)
```

### 1.2 Generate Import Template (Optional)

If you're starting from scratch, generate a template to understand required fields:

```{r}
# Generate a template with examples
template <- get_plot_metadata_template(
  template_type = "permanent_plot",  # or "transect", "minimal", "full"
  with_examples = TRUE
)

# View template structure
View(template)

# Export to Excel for filling
export_plot_template(
  file_path = "my_plots_template.xlsx",
  template_type = "permanent_plot",
  with_examples = TRUE
)

# See what columns are available
print_template_info("permanent_plot")
```

**Template includes**:
- **Required**: `plot_name`, `method`, `country`
- **Location**: `ddlat`, `ddlon`, `elevation`, `locality_name`
- **Dates**: `date_y`, `date_m`, `date_d`
- **People**: `team_leader`, `principal_investigator`, `data_manager`
- **Optional**: Plot characteristics and other subplot features

### 1.3 Map Your Data Columns

Map your data columns to the database schema:

```{r}
# Load your data (from Excel, CSV, etc.)
my_plot_data <- readxl::read_excel("my_plots.xlsx")
# or
my_plot_data <- read.csv("my_plots.csv")

# Get import configuration
config <- get_import_column_routing(import_type = "plots")

# Map columns automatically (with fuzzy matching)
mapping_result <- map_user_columns(
  data = my_plot_data,
  config = config,
  similarity_threshold = 0.6,
  interactive = TRUE  # Set FALSE for non-interactive
)

# Review mappings
print_mapping_summary(mapping_result)
```

**Column Mapping Features**:
- ✅ Automatic exact matching
- ✅ Synonym recognition (e.g., `PI` → `principal_investigator`)
- ✅ Fuzzy string matching for typos
- ✅ Interactive selection for unclear mappings

### 1.4 Validate Data

Validate your data before importing:

```{r}
# Validate plot metadata
validation <- validate_plot_metadata(
  data = my_plot_data,
  column_mappings = mapping_result$mappings,
  config = config,
  interactive = TRUE,  # Allow interactive fixing
  fix_on_fly = TRUE    # Fix issues during validation
)

# View validation results
print_validation_results(validation)

# Check if valid
if (!validation$valid) {
  cat("❌ Validation failed! Fix errors before importing.\n")
  # View detailed errors
  View(validation$errors)
  stop("Data validation failed")
}
```

**Validation Checks**:
1. ✅ Required fields present (`plot_name`, `method`, `country`)
2. ✅ Lookup values valid (method, country)
3. ✅ Column types correct (numeric, character)
4. ✅ Value ranges valid (coordinates, dates)
5. ✅ Unique constraints (no duplicate plot names)

**Interactive Fixing**: The validation function can automatically fix lookup mismatches using fuzzy matching if `interactive = TRUE`.

### 1.5 Dry Run (Preview Import)

Always preview before committing:

```{r}
# DRY RUN - preview without committing
preview <- import_plot_metadata(
  data = my_plot_data,
  column_mappings = mapping_result$mappings,
  validation = validation,
  config = config,
  dry_run = TRUE,  # IMPORTANT: Preview first!
  con = con
)

# Review what would be imported
print_import_result(preview)
```

**Dry Run Output**:
```
═══ Dry Run: Preview Import (No Changes Will Be Made) ═══
ℹ Plots to import: 3
ℹ Importing as user: your_username

─── Step 1: Linking methods ───
✓ Would link 2 unique methods

─── Step 2: Linking countries ───
✓ Would link 2 unique countries

─── Step 3: Processing subplot features ───
✓ Found 4 subplot feature column(s): team_leader, principal_investigator, ...

─── Step 4: Preparing plot data ───
✓ Prepared 3 rows with 15 columns

─── Step 5: Preview - Would Insert Into data_liste_plots ───
Data preview (first 3 rows):
  plot_name     method  id_method id_country ddlat  ddlon
1 Plot-A        1ha-IRD    1          2      -0.50  11.50
2 Plot-B        1ha-IRD    1          2      -0.52  11.48
3 Plot-C        0.5ha       5          1       3.20  13.50

─── Step 6: Preview subplot features ───
ℹ Would insert 3 team_leader records (people)
ℹ Would insert 3 principal_investigator records (people)

✓ Dry run completed - no changes made
```

### 1.6 Actual Import

If dry run looks good, proceed with actual import:

```{r}
# ACTUAL IMPORT
result <- import_plot_metadata(
  data = my_plot_data,
  column_mappings = mapping_result$mappings,
  validation = validation,
  config = config,
  dry_run = FALSE,  # Actually import
  interactive = TRUE,
  con = con
)

# Check result
if (result$success) {
  cat("✅ Successfully imported", result$n_plots, "plots!\n")
  cat("Plot names:", paste(result$plot_names, collapse = ", "), "\n")
} else {
  cat("❌ Import failed:", result$message, "\n")
}
```

**Import Process**:
1. Transaction started (all-or-nothing)
2. Methods linked to `methodslist` table
3. Countries linked to `table_countries` table
4. People linked to `table_colnam` table (interactive matching)
5. Plot data inserted into `data_liste_plots`
6. Subplot features inserted into `data_liste_sub_plots`
7. Transaction committed (or rolled back on error)

### 1.7 Row-Level Security (RLS) - Important!

**After import, you may not have access to your plots yet!**

The function generates admin access code:

```{r}
# Admin code is returned in the result
cat(result$admin_code)

# Or save to file to send to admin
writeLines(result$admin_code, "admin_access_request.R")
```

**Send this code to your database administrator** to grant you access:

```r
# ══════════════════════════════════════════════════════════════════════
# ROW-LEVEL SECURITY: Grant Access to User
# ══════════════════════════════════════════════════════════════════════
# User: your_username
# Plots: Plot-A, Plot-B, Plot-C

library(CafriplotsR)
con <- call.mydb()  # Admin credentials

# Get plot IDs
plot_ids <- DBI::dbGetQuery(con,
  "SELECT id_liste_plots FROM data_liste_plots
   WHERE plot_name IN ('Plot-A', 'Plot-B', 'Plot-C')")$id_liste_plots

# Grant access
define_user_policy(
  con = con,
  user = "your_username",
  ids = plot_ids,
  table = "data_liste_plots",
  operations = c("SELECT", "UPDATE"),
  drop_existing = TRUE
)
```

### 1.8 Verify Import

After admin grants access, verify plots were created:

```{r}
# Query imported plots
imported_plots <- query_plots(
  plot_name = c("Plot-A", "Plot-B", "Plot-C"),
  exact_match = TRUE,
  show_multiple_census = TRUE,
  con = con
)

# Check metadata
print(imported_plots$metadata)

# Check census features
if (!is.null(imported_plots$census_features)) {
  print(imported_plots$census_features)
}
```

**Important**: Ensure all plots are successfully imported before proceeding to individual data import.

---

## Step 2: Import Individual Tree Data

Once plots exist in the database, you can import individual tree measurements and traits.

### 2.1 Understanding the Data Structure

Individual data can be organized in two ways:

#### Option A: Flat Table (All Data in One Sheet)

```
plot_name | tag | species      | idtax | dbh  | height | wood_density
----------|-----|--------------|-------|------|--------|-------------
Plot-A    | 1   | Sp1          | 123   | 35.2 | 18.5   | 0.65
Plot-A    | 2   | Sp2          | 456   | 42.1 | 22.0   | 0.58
```

#### Option B: Two-Table Structure (Individuals + Features)

**Individuals Sheet** (core tree identification):
```
plot_name | tag | species | idtax | herbarium_code
----------|-----|---------|-------|---------------
Plot-A    | 1   | Sp1     | 123   | IRD001
Plot-A    | 2   | Sp2     | 456   | IRD002
```

**Features Sheet** (measurements and traits):
```
plot_name | tag | dbh  | height | wood_density | census_date
----------|-----|------|--------|--------------|------------
Plot-A    | 1   | 35.2 | 18.5   | 0.65         | 2020-01-15
Plot-A    | 2   | 42.1 | 22.0   | 0.58         | 2020-01-15
```

### 2.2 Generate Import Template

Start by generating a template to understand required columns:

```{r}
# Generate template with guidance
get_individual_template(output_file = "my_individual_template.xlsx")

# This creates an Excel file with:
# - Individuals sheet with required columns
# - Features sheet for trait measurements
# - Notes sheet with detailed guidance

# View template info
print_individual_template_info()
```

**Template Structure**:

- **Required columns**: `plot_name`, `idtax_n` (taxonomy ID)
- **Recommended columns**: `tag` (tree identifier), `original_tax_name`
- **Optional columns**: `herbarium_nbe_type`, `herbarium_nbe_char`, `multi_tiges_id`
- **Features**: Any trait from `traits_list()` (e.g., `stem_diameter`, `tree_height`, `wood_density`)

### 2.3 Prepare Your Data

Load your tree inventory data:

```{r}
# Option 1: Load from Excel (flat table)
my_data <- readxl::read_excel("my_tree_inventory.xlsx")

# Option 2: Load from Excel (two-table structure)
individuals_raw <- readxl::read_excel("my_tree_inventory.xlsx", sheet = "Individuals")
features_raw <- readxl::read_excel("my_tree_inventory.xlsx", sheet = "Features")

# Option 3: Load from CSV
my_data <- read.csv("my_tree_inventory.csv")
```

### 2.4 Column Mapping

Map your data columns to database schema using the interactive mapper:

```{r}
# INTERACTIVE MODE (recommended for first-time users)
mapped <- map_individual_columns(
  data = my_data,
  interactive = TRUE,
  con = con
)

# NON-INTERACTIVE MODE (with synonyms for faster mapping)
mapped <- map_individual_columns(
  data = my_data,
  interactive = FALSE,
  use_synonyms = TRUE,
  con = con
)

# For two-table structure
mapped <- map_individual_columns(
  individuals = individuals_raw,
  features = features_raw,
  interactive = TRUE,
  con = con
)
```

**Column Mapping Process**:

1. **Automatic matching**: The function fuzzy-matches your column names to database columns
2. **Feature classification**: You'll be asked to classify unmapped columns:
   - **YES** - It's a trait/feature (e.g., DBH, height, leaf area)
   - **NO** - It's identification data (e.g., plot name, tree tag, species)
   - **SKIP** - Ignore this column
3. **Manual selection**: Choose from suggested matches or skip columns
4. **Summary**: Review the mapping before proceeding

**Example Interactive Session**:
```
Column: 'TreeID'
Is this a feature/trait measurement? (YES/NO/SKIP): NO
Select the best match for 'TreeID':
  1. tag (Individual tree identifier)
  2. id_n (Database ID)
  3. herbarium_nbe_char (Herbarium code)
  [0 to skip]
Your choice: 1
✓ Mapped TreeID → tag
```

**Result**:
```{r}
# Mapped data structure
str(mapped)
# $individuals - Core tree data with standardized column names
# $features - Trait measurements (if features were present)
# $mapping - Record of column name mappings
```

### 2.5 Validation

Validate your data before importing to catch errors early:

```{r}
# Validate mapped data
validation <- validate_individual_data(
  individuals_data = mapped$individuals,
  features_data = mapped$features,
  method = "1ha-IRD",  # Your plot method
  con = con
)

# View validation results
print_individual_validation_results(validation)
```

**Validation Checks**:

1. ✓ **Required columns present** (`plot_name`, `idtax_n`)
2. ✓ **Plot existence** - All plots exist in database
3. ✓ **Taxonomy validation** - All `idtax_n` values are valid
4. ✓ **Tag uniqueness** - No duplicate tags within plots
5. ✓ **Tag conflicts** - No conflicts with existing database tags
6. ✓ **Trait value validity** - Numeric traits are numeric, categorical values are valid
7. ✓ **Feature linkage** - Features link correctly to individuals

**Validation Output**:
```
═══ VALIDATION SUMMARY ═══
Status: ✓ VALID (Ready to import)

Data Summary:
  • 11,158 individuals across 32 plots
  • 4,523 feature records
  • 0 critical errors
  • 2 warnings

Warnings:
  ⚠ Missing optional column: herbarium_nbe_char
  ⚠ 15 individuals missing tree_height measurements
```

**If validation fails**:
```{r}
if (!validation$valid) {
  # View detailed errors
  print_individual_validation_results(validation)

  # Fix issues in your source data and re-map
  # Common issues:
  # - Plot names not matching database (use exact_match in query_plots)
  # - Invalid taxonomy IDs (re-run taxonomic matching)
  # - Duplicate tags (check your source data)

  stop("Fix validation errors before importing")
}
```

### 2.6 Dry Run (Preview Import)

Always preview the import before committing to the database:

```{r}
# DRY RUN - Preview without committing
preview <- import_individual_data(
  individuals_data = validation$cleaned_data$individuals,
  features_data = validation$cleaned_data$features,
  validation = validation,
  method = "1ha-IRD",
  dry_run = TRUE,
  con = con
)

# Review what would be imported
print(preview)
```

**Dry Run Output**:
```
═══ Dry Run: Preview Import (No Changes Will Be Made) ═══
ℹ Individuals to import: 11,158
ℹ Feature records to import: 4,523
ℹ Importing as user: your_username

─── Step 1: Linking individuals to plots ───
✓ Linked 11,158 individuals to 32 plots

─── Step 2: Preparing individual data ───
✓ Prepared 11,158 individual records
ℹ Columns: id_table_liste_plots_n, tag, idtax_n, original_tax_name

─── Step 3: Preview - Would Insert Into data_individuals ───
Data preview (first 3 rows):
  id_table_liste_plots_n tag idtax_n original_tax_name
1                    145   1    1234      Sp1
2                    145   2    5678      Sp2
3                    145   3    9012      Sp3

─── Step 4: Preview individual features ───
ℹ Would insert 4,523 feature records
Features preview (first 3 rows):
  id_data_individuals traitid traitvalue
1                   1       5       35.2
2                   1       8       18.5
3                   2       5       42.1

✓ Dry run completed - no changes made
ℹ Run with dry_run = FALSE to actually import
```

### 2.7 Actual Import

If the dry run looks good, proceed with the actual import:

```{r}
# ACTUAL IMPORT with transaction support
result <- import_individual_data(
  individuals_data = validation$cleaned_data$individuals,
  features_data = validation$cleaned_data$features,
  validation = validation,
  method = "1ha-IRD",
  dry_run = FALSE,
  ask_confirmation = TRUE,  # Prompt for confirmation
  con = con
)

# Check result
if (result$success) {
  cat("✓ Successfully imported:\n")
  cat("  -", result$n_individuals, "individuals\n")
  cat("  -", result$n_features, "feature records\n")
  cat("  - Affected plots:", paste(result$plot_names, collapse = ", "), "\n")
} else {
  cat("✗ Import failed:", result$message, "\n")
}
```

**Import Process**:

1. **User confirmation** - You'll be asked to confirm before proceeding
2. **Transaction started** - All-or-nothing import begins
3. **Link to plots** - Individuals are linked to plot IDs
4. **Insert individuals** - Core data inserted into `data_individuals`
5. **Insert features** - Traits inserted into `data_traits_measures`
6. **Transaction committed** - Changes are saved
7. **Automatic rollback** - If any error occurs, all changes are reverted

**Import Output**:
```
═══ Importing Individual Data ═══
ℹ Individuals to import: 11,158
ℹ Feature records to import: 4,523
ℹ Importing as user: your_username

Proceed with import? (yes/no): yes

ℹ Transaction started
✓ Linked 11,158 individuals to 32 plots
✓ Prepared 11,158 individual records
✓ 11,158 individuals inserted
✓ 4,523 feature records inserted
✓ Transaction committed successfully

═══════════════════════════════════════
✓ Import completed successfully!

ℹ Imported 11,158 individuals
ℹ Imported 4,523 feature records
ℹ Affected plots: Plot-A, Plot-B, Plot-C, ...
═══════════════════════════════════════
```

### 2.8 Verify Import

After importing, verify the data was correctly added:

```{r}
# Query the imported data
imported_data <- query_plots(
  plot_name = c("Plot-A", "Plot-B", "Plot-C"),
  exact_match = TRUE,
  extract_individuals = TRUE,
  output_style = "permanent_plot",
  con = con
)

# Check individuals
head(imported_data$individuals)

# Check metadata
imported_data$metadata

# Verify counts
cat("Imported individuals per plot:\n")
table(imported_data$individuals$plot_name)
```

---

## Complete Example Workflow

Here's a complete end-to-end example:

```{r}
library(CafriplotsR)

# 1. Connect
con <- call.mydb()

# 2. Load raw data
raw_data <- readxl::read_excel("my_field_data.xlsx")

# 3. Ensure plots exist (import plot metadata first if needed)
existing_plots <- query_plots(
  plot_name = unique(raw_data$PlotName),
  exact_match = TRUE,
  con = con
)

if (nrow(existing_plots) == 0) {
  stop("Plots not found! Import plot metadata first.")
}

# 4. Map columns
mapped <- map_individual_columns(
  data = raw_data,
  interactive = TRUE,
  con = con
)

# 5. Validate
validation <- validate_individual_data(
  individuals_data = mapped$individuals,
  features_data = mapped$features,
  method = "1ha-IRD",
  con = con
)

# 6. Check validation
if (!validation$valid) {
  print_individual_validation_results(validation)
  stop("Fix validation errors first!")
}

# 7. Dry run
preview <- import_individual_data(
  individuals_data = validation$cleaned_data$individuals,
  features_data = validation$cleaned_data$features,
  validation = validation,
  dry_run = TRUE,
  con = con
)

# 8. Actual import
result <- import_individual_data(
  individuals_data = validation$cleaned_data$individuals,
  features_data = validation$cleaned_data$features,
  validation = validation,
  dry_run = FALSE,
  con = con
)

# 9. Verify
if (result$success) {
  cat("✓ Import successful!\n")

  # Query imported data
  verification <- query_plots(
    plot_name = result$plot_names,
    exact_match = TRUE,
    extract_individuals = TRUE,
    con = con
  )

  cat("Verified:", nrow(verification$individuals), "individuals imported\n")
} else {
  cat("✗ Import failed:", result$message, "\n")
}

# 10. Cleanup
DBI::dbDisconnect(con)
```

---

## Common Issues and Solutions

### Issue: "Plot not found in database"

**Solution**: Import plot metadata first before importing individuals.

```{r}
# Check if plots exist
query_plots(plot_name = "Plot-A", exact_match = TRUE, con = con)

# If not found, import plot metadata first
```

### Issue: "Invalid taxonomy ID"

**Solution**: Standardize taxonomic names using the taxonomic matching app.

```{r}
# Launch taxonomic matching app
launch_taxonomic_match_app(data = my_data, name_column = "species")

# Or use programmatic matching
matched <- match_taxonomic_names(my_data, name_column = "species", con_taxa = con_taxa)
```

### Issue: "Duplicate tags within plot"

**Solution**: Ensure tags are unique within each plot, or let the function auto-generate tags.

```{r}
# Remove tag column to auto-generate sequential tags
my_data$tag <- NULL

# Or manually fix duplicates
my_data <- my_data %>%
  group_by(plot_name) %>%
  mutate(tag = row_number()) %>%
  ungroup()
```

### Issue: "Column mapping unclear"

**Solution**: Use interactive mode and carefully read the prompts. When classifying columns:
- **Features** = measurements/traits (DBH, height, density)
- **Individuals** = identification (plot, tag, species)

### Issue: "Transaction rollback on error"

**Solution**: Check the error message carefully. Common causes:
- Missing required columns
- Invalid foreign key references
- Data type mismatches

All changes are automatically rolled back on error, so your database remains consistent.

---

## Best Practices

1. **Always validate first** - Use `validate_individual_data()` before importing
2. **Use dry run** - Preview with `dry_run = TRUE` before actual import
3. **Standardize taxonomy** - Match taxonomic names before importing individuals
4. **Import plots first** - Ensure plot metadata exists before adding individuals
5. **Use transactions** - The import functions use transactions for safety
6. **Keep backups** - Maintain backups of your database
7. **Document changes** - Keep track of what data was imported and when
8. **Check permissions** - Ensure you have write permissions to the database

---

## Advanced: Programmatic Column Mapping

For repeated imports with consistent data formats, you can use synonyms to automate mapping:

```{r}
# Define custom synonyms
custom_synonyms <- list(
  tag = c("TreeID", "Tree_Number", "Individual_ID"),
  stem_diameter = c("DBH", "Diameter", "dbh_cm"),
  tree_height = c("Height", "H", "height_m")
)

# Map with custom synonyms
mapped <- map_individual_columns(
  data = my_data,
  interactive = FALSE,
  use_synonyms = TRUE,
  custom_synonyms = custom_synonyms,
  con = con
)
```

---

## Additional Resources

- **Taxonomic matching**: See `vignette("taxonomic-app")`
- **Querying data**: See `vignette("using-query-plots")`
- **Template functions**: `?get_individual_template`, `?print_individual_template_info`
- **Validation functions**: `?validate_individual_data`, `?print_individual_validation_results`
- **Import functions**: `?import_individual_data`
- **Column mapping**: `?map_individual_columns`

---

## Summary

The complete workflow for importing plot data:

1. ✅ **Import plot metadata** (location, census dates, etc.)
2. ✅ **Standardize taxonomy** using `launch_taxonomic_match_app()`
3. ✅ **Map columns** using `map_individual_columns()`
4. ✅ **Validate data** using `validate_individual_data()`
5. ✅ **Preview import** with `dry_run = TRUE`
6. ✅ **Import data** with `dry_run = FALSE`
7. ✅ **Verify import** using `query_plots()`

All imports use **transaction-based safety** with automatic rollback on errors, ensuring your database remains consistent even if something goes wrong.
