---
title: "Importing Plot Data into the Database"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Importing Plot Data into the Database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  echo = TRUE
)
```

## Overview

This vignette demonstrates how to import new plot data into the database. The package provides **two approaches**:

1. **✨ Interactive Shiny App (Recommended)** - User-friendly guided workflow with visual preview and validation
2. **💻 Programmatic Workflow (Advanced)** - R code for automation and custom workflows

We **strongly recommend using the Shiny App** for most users, especially when:
- Importing data for the first time
- Working with new data formats
- You want visual confirmation before importing
- You prefer a guided step-by-step process

The programmatic workflow is for advanced users who need automation or custom integration.

## Prerequisites

Before importing data, ensure you have:

1. **Database credentials** configured with write permissions
2. **Taxonomic names standardized** using `launch_taxonomic_match_app()` (see taxonomic-app vignette)
3. **Plot coordinates and metadata** in Excel or CSV format
4. **Individual tree data** (if importing tree measurements)

```{r}
library(CafriplotsR)

# Connect to database (for verification)
con <- connect_cafri()$main
```

---

## Import Using the Shiny App (Recommended)

The **Import Wizard** provides a comprehensive, visual interface for importing plot metadata with built-in validation, preview, and error handling.

### Launch the Import Wizard

```{r}
# Launch the Import Wizard
launch_import_wizard()

# Or specify language
launch_import_wizard(language = "fr")  # French interface
```

### Step-by-Step Guide

The Import Wizard guides you through 7 steps:

#### **Step 1: Choose Import Type**

Select what you want to import:
- **Plot Metadata** - Plot locations, census dates, investigators
- **Individual Trees** - Tree measurements and traits (coming soon)

#### **Step 2: Upload Data**

- **Upload your file** (Excel `.xlsx` or CSV)
- **Or download a template** if starting from scratch
- Preview your data to ensure it loaded correctly

**Supported formats:**
- Excel files (`.xlsx`, `.xls`)
- CSV files (`,` or `;` delimited)
- Multiple sheets (you can select which sheet to import)

#### **Step 3: Map Columns**

The wizard automatically maps your column names to the database schema.

**Required vs Optional Columns:**

At the top of this step, you'll see clearly which columns are **mandatory** vs **recommended**:

- **📌 Required Columns (Mandatory)** - Highlighted in yellow/gold
  - For **Plot Metadata**: `plot_name`, `method`, `country`
  - For **Individual Trees**: `plot_name`, `idtax_n`, `original_tax_name`
  - You **cannot proceed** until all required columns are mapped

- **💡 Recommended Columns (Optional)** - Highlighted in blue
  - For plots: `ddlat`, `ddlon`, `date_y`, `locality_name`, etc.
  - These enhance data quality but are not mandatory
  - You can skip these if you don't have the data

**Automatic Mapping:**

- **✅ Exact matches** - Highlighted in green
- **🔍 Fuzzy matches** - Suggested with confidence scores
- **❓ Unmapped columns** - You can manually select or skip

**Smart features:**
- Recognizes synonyms (e.g., `PI` → `principal_investigator`, `dbh` → `stem_diameter`)
- Provides column descriptions on hover
- Allows you to skip unwanted columns
- Shows sample values from your data for context
- Remembers your choices when navigating back

**Validation:**
- The wizard validates that all required columns are mapped
- Shows clear error messages if required columns are missing
- Prevents proceeding to the next step until validation passes

#### **Step 4: Match Lookup Values**

For columns that reference lookup tables (method, country, people), the wizard:

- Shows exact matches in green
- Highlights values needing matching
- Provides interactive fuzzy matching with similarity scores
- Lets you search and select correct values

**Example:**
```
Your value: "Cameroun"
Suggestions:
  ✓ CAMEROON (95% match)
  • CAMEROUN (exact but not in database)
  • Gabon (20% match)
```

#### **Step 5: Validate Data**

Comprehensive validation checks:

✅ **Required fields** - plot_name, method, country
✅ **Data types** - Numeric columns contain numbers
✅ **Value ranges** - Coordinates within valid ranges
✅ **Lookup values** - All references exist in database
✅ **Unique constraints** - No duplicate plot names
✅ **Duplicate detection** - Warns about potential re-imports

**UTM Coordinate Detection:**
If your coordinates are in UTM format (values > 1000), the wizard:
- Detects them automatically
- Shows a specific error message
- Provides conversion tool in the preview step

#### **Step 6: Preview Data**

Visual preview of your data **before** importing:

- **Data table** with enriched lookup names (not IDs)
- **Interactive map** showing plot locations
  - Auto-zoom to your plots
  - Marker clustering for many plots
  - Click markers for plot details
  - Warnings for invalid/unusual coordinates
- **UTM conversion tool** (if UTM coordinates detected)
  - Input: UTM zone and hemisphere
  - Output: Converted WGS84 coordinates
  - Updates map in real-time
- **Download options** - Excel or CSV with enriched data

**Map Features:**
- 🗺️ Verify plot locations visually
- ⚠️ Catch reversed latitude/longitude
- 📍 Check for coordinate system issues
- 💡 Helpful tips if plots aren't where expected

#### **Step 7: Execute Import**

Final step to import into the database:

- **Dry Run** mode to test without changes
- **Live progress tracking** during import
- **Transaction safety** - All-or-nothing import
- **Admin access code** generated for row-level security
- **Download results** as R script or CSV

**After Import:**
The wizard provides an **admin access code** that grants you permission to access your newly imported plots. Send this code to your database administrator.

### Complete Example Session

```{r}
# 1. Launch wizard
launch_import_wizard()

# 2. Follow the 7-step guided process:
#    - Choose "Plot Metadata"
#    - Upload your Excel file
#    - Review automatic column mapping
#    - Match lookup values interactively
#    - Validate data (fix any errors)
#    - Preview data and map
#    - Execute import (try dry run first!)

# 3. Copy admin access code and send to admin

# 4. After admin grants access, verify:
con <- connect_cafri()$main
my_plots <- query_plots(
  plot_name = c("Plot-A", "Plot-B"),
  exact_match = TRUE,
  con = con
)
```

### Import Wizard Features

**Validation & Error Prevention:**
- ✅ Duplicate plot detection (method + country + coordinates)
- ✅ UTM coordinate detection and conversion
- ✅ Interactive map preview
- ✅ Lookup value matching with fuzzy search
- ✅ Real-time validation feedback
- ✅ Reversed lat/lon detection

**User-Friendly Design:**
- 📊 Visual data preview
- 🗺️ Interactive maps
- 🔍 Intelligent column mapping
- ⚡ Instant feedback
- 💾 Download options at every step
- 🌍 Bilingual (English/French)

**Safety Features:**
- 🔒 Dry run mode
- ↩️ Transaction rollback on errors
- ✅ Step-by-step validation
- 📋 Comprehensive error messages

---

## When to Use Programmatic Workflow

The programmatic workflow is designed for **advanced users** who need:

- **Automation** - Batch processing of multiple files
- **Custom validation** - Additional business logic
- **Integration** - Part of a larger data pipeline
- **Scripting** - Reproducible import workflows
- **Performance** - Import very large datasets efficiently

**For most users, the Shiny App is recommended** as it provides:
- Better error messages
- Visual confirmation
- Interactive problem-solving
- No coding required

---

## Programmatic Import (Advanced Users)

<details>
<summary><strong>Click to expand programmatic workflow documentation</strong></summary>

This section provides the complete R code workflow for importing data programmatically.

### Step 1: Import Plot Metadata

#### 1.1 Prepare Plot Data

Plot data should include essential metadata:

```{r}
# Example: Prepare plot metadata
plot_metadata <- data.frame(
  plot_name = c("Plot-A", "Plot-B", "Plot-C"),
  country = c("Gabon", "Gabon", "Cameroon"),
  locality_name = c("Lopé NP", "Lopé NP", "Dja Reserve"),
  ddlat = c(-0.5, -0.52, 3.2),
  ddlon = c(11.5, 11.48, 13.5),
  plot_area = c(1, 1, 0.5),  # hectares
  date_census_1 = as.Date(c("2020-01-15", "2020-02-10", "2019-11-20")),
  team_leader_1 = c("John Doe", "Jane Smith", "Bob Wilson"),
  principal_investigator_1 = c("Dr. Smith", "Dr. Smith", "Dr. Jones"),
  method = c("1ha-IRD", "1ha-IRD", "0.5ha-custom"),
  subplot_shape = c("square", "square", "rectangle"),
  stringsAsFactors = FALSE
)
```

#### 1.2 Generate Import Template (Optional)

```{r}
# Generate a template with examples
template <- get_plot_metadata_template(
  template_type = "permanent_plot",
  with_examples = TRUE
)

# Export to Excel
export_plot_template(
  file_path = "my_plots_template.xlsx",
  template_type = "permanent_plot",
  with_examples = TRUE
)

# See available columns
print_template_info("permanent_plot")
```

#### 1.3 Map Your Data Columns

```{r}
# Load your data
my_plot_data <- readxl::read_excel("my_plots.xlsx")

# Get import configuration
config <- get_import_column_routing(import_type = "plots")

# Map columns automatically
mapping_result <- map_user_columns(
  data = my_plot_data,
  config = config,
  similarity_threshold = 0.6,
  interactive = TRUE
)

# Review mappings
print_mapping_summary(mapping_result)
```

#### 1.4 Validate Data

```{r}
# Validate plot metadata
validation <- validate_plot_metadata(
  data = my_plot_data,
  column_mappings = mapping_result$mappings,
  config = config,
  interactive = TRUE,
  fix_on_fly = TRUE
)

# View results
print_validation_results(validation)

# Check validity
if (!validation$valid) {
  View(validation$errors)
  stop("Fix validation errors before importing")
}
```

#### 1.5 Dry Run (Preview)

```{r}
# Preview without committing
preview <- import_plot_metadata(
  data = my_plot_data,
  column_mappings = mapping_result$mappings,
  validation = validation,
  config = config,
  dry_run = TRUE,
  con = con
)

print_import_result(preview)
```

#### 1.6 Actual Import

```{r}
# Import to database
result <- import_plot_metadata(
  data = my_plot_data,
  column_mappings = mapping_result$mappings,
  validation = validation,
  config = config,
  dry_run = FALSE,
  interactive = TRUE,
  con = con
)

if (result$success) {
  cat("✅ Successfully imported", result$n_plots, "plots!\n")

  # Save admin access code
  writeLines(result$admin_code, "admin_access_request.R")
}
```

#### 1.7 Row-Level Security Access

Send the generated admin code to your database administrator:

```{r}
# Admin runs this code to grant access
library(CafriplotsR)
con <- connect_cafri()$main

plot_ids <- DBI::dbGetQuery(con,
  "SELECT id_liste_plots FROM data_liste_plots
   WHERE plot_name IN ('Plot-A', 'Plot-B', 'Plot-C')")$id_liste_plots

define_user_policy(
  con = con,
  user = "your_username",
  ids = plot_ids,
  table = "data_liste_plots",
  operations = c("SELECT", "UPDATE"),
  drop_existing = TRUE
)
```

#### 1.8 Verify Import

```{r}
# After admin grants access
imported_plots <- query_plots(
  plot_name = c("Plot-A", "Plot-B", "Plot-C"),
  exact_match = TRUE,
  show_multiple_census = TRUE,
  con = con
)

print(imported_plots$metadata)
```

---

### Step 2: Import Individual Tree Data

Once plots exist, import individual tree measurements.

#### 2.1 Data Structure

**Option A: Flat Table**
```
plot_name | tag | idtax | dbh  | height
----------|-----|-------|------|-------
Plot-A    | 1   | 123   | 35.2 | 18.5
Plot-A    | 2   | 456   | 42.1 | 22.0
```

**Option B: Two Tables**
- **Individuals**: plot_name, tag, idtax, species
- **Features**: plot_name, tag, dbh, height, wood_density

#### 2.2 Generate Template

```{r}
# Generate template
get_individual_template(output_file = "tree_template.xlsx")

# View template info
print_individual_template_info()
```

#### 2.3 Load and Map Data

```{r}
# Load data
my_data <- readxl::read_excel("my_trees.xlsx")

# Map columns interactively
mapped <- map_individual_columns(
  data = my_data,
  interactive = TRUE,
  con = con
)
```

#### 2.4 Validate

```{r}
# Validate
validation <- validate_individual_data(
  individuals_data = mapped$individuals,
  features_data = mapped$features,
  method = "1ha-IRD",
  con = con
)

# Check results
print_individual_validation_results(validation)

if (!validation$valid) {
  stop("Fix validation errors first!")
}
```

#### 2.5 Import

```{r}
# Dry run
preview <- import_individual_data(
  individuals_data = validation$cleaned_data$individuals,
  features_data = validation$cleaned_data$features,
  validation = validation,
  dry_run = TRUE,
  con = con
)

# Actual import
result <- import_individual_data(
  individuals_data = validation$cleaned_data$individuals,
  features_data = validation$cleaned_data$features,
  validation = validation,
  dry_run = FALSE,
  con = con
)

if (result$success) {
  cat("✓ Imported", result$n_individuals, "individuals\n")
}
```

#### 2.6 Verify

```{r}
# Query imported data
verification <- query_plots(
  plot_name = result$plot_names,
  exact_match = TRUE,
  extract_individuals = TRUE,
  con = con
)

head(verification$individuals)
```

---

### Complete Programmatic Example

```{r}
library(CafriplotsR)

# 1. Connect
con <- connect_cafri()$main

# 2. Load data
raw_data <- readxl::read_excel("my_field_data.xlsx")

# 3. Verify plots exist
existing_plots <- query_plots(
  plot_name = unique(raw_data$PlotName),
  exact_match = TRUE,
  con = con
)

if (nrow(existing_plots) == 0) {
  stop("Import plot metadata first!")
}

# 4. Map columns
mapped <- map_individual_columns(
  data = raw_data,
  interactive = TRUE,
  con = con
)

# 5. Validate
validation <- validate_individual_data(
  individuals_data = mapped$individuals,
  features_data = mapped$features,
  method = "1ha-IRD",
  con = con
)

if (!validation$valid) {
  print_individual_validation_results(validation)
  stop("Fix errors first!")
}

# 6. Dry run
preview <- import_individual_data(
  individuals_data = validation$cleaned_data$individuals,
  features_data = validation$cleaned_data$features,
  validation = validation,
  dry_run = TRUE,
  con = con
)

# 7. Import
result <- import_individual_data(
  individuals_data = validation$cleaned_data$individuals,
  features_data = validation$cleaned_data$features,
  validation = validation,
  dry_run = FALSE,
  con = con
)

# 8. Verify
if (result$success) {
  verification <- query_plots(
    plot_name = result$plot_names,
    exact_match = TRUE,
    extract_individuals = TRUE,
    con = con
  )

  cat("✓ Verified:", nrow(verification$individuals), "individuals\n")
}

# 9. Cleanup
DBI::dbDisconnect(con)
```

---

### Common Issues and Solutions

**Issue: "Plot not found in database"**
```{r}
# Solution: Import plot metadata first
query_plots(plot_name = "Plot-A", exact_match = TRUE, con = con)
```

**Issue: "Invalid taxonomy ID"**
```{r}
# Solution: Use taxonomic matching app
launch_taxonomic_match_app(data = my_data, name_column = "species")
```

**Issue: "Duplicate tags within plot"**
```{r}
# Solution: Auto-generate tags
my_data$tag <- NULL  # Remove tag column

# Or fix manually
my_data <- my_data %>%
  group_by(plot_name) %>%
  mutate(tag = row_number()) %>%
  ungroup()
```

**Issue: "Transaction rollback on error"**
- Check error message carefully
- All changes rolled back automatically
- Database remains consistent

---

### Advanced: Custom Column Mapping

```{r}
# Define custom synonyms for repeated imports
custom_synonyms <- list(
  tag = c("TreeID", "Tree_Number", "Individual_ID"),
  stem_diameter = c("DBH", "Diameter", "dbh_cm"),
  tree_height = c("Height", "H", "height_m")
)

# Map with custom synonyms (non-interactive)
mapped <- map_individual_columns(
  data = my_data,
  interactive = FALSE,
  use_synonyms = TRUE,
  custom_synonyms = custom_synonyms,
  con = con
)
```

---

### Best Practices

1. ✅ **Always validate first** - Catch errors early
2. ✅ **Use dry run** - Preview before committing
3. ✅ **Standardize taxonomy** - Match names before importing
4. ✅ **Import plots first** - Metadata before individuals
5. ✅ **Use transactions** - Automatic rollback on errors
6. ✅ **Keep backups** - Regular database backups
7. ✅ **Document changes** - Track what was imported
8. ✅ **Check permissions** - Ensure write access

</details>

---

## Summary

### Recommended Workflow (Shiny App)

1. ✨ **Launch Import Wizard** - `launch_import_wizard()`
2. 📤 **Upload your data** - Excel or CSV
3. 🔗 **Map columns** - Automatic with fuzzy matching
4. 🔍 **Match lookups** - Interactive value matching
5. ✅ **Validate** - Comprehensive checks + duplicate detection
6. 🗺️ **Preview** - Visual confirmation with map
7. 💾 **Import** - Execute with transaction safety

### Advanced Workflow (Programmatic)

For automation and custom workflows:
- `map_user_columns()` - Column mapping
- `validate_plot_metadata()` - Data validation
- `import_plot_metadata()` - Import plots
- `map_individual_columns()` - Tree data mapping
- `validate_individual_data()` - Tree validation
- `import_individual_data()` - Import trees

---

## Additional Resources

- **Taxonomic matching**: `vignette("taxonomic-app")`
- **Querying data**: `vignette("using-query-plots")`
- **Template functions**: `?get_plot_metadata_template`, `?get_individual_template`
- **Validation**: `?validate_plot_metadata`, `?validate_individual_data`
- **Import Wizard**: `?launch_import_wizard`
