---
title: "Managing table_idtax Materialized View (Advanced)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Managing table_idtax Materialized View (Advanced)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

**Note: This vignette is for advanced users with database management permissions only.**

The `table_idtax` is a PostgreSQL materialized view that resolves taxonomic synonyms by linking `idtax` (any taxon ID) to `idtax_good` (the accepted taxon ID). It is based on data from the `table_taxa` table in the taxa database.

## Why a Materialized View?

Previously, `table_idtax` was a regular table that required manual updates by the database administrator. As a materialized view:

- **Performance**: Query results are pre-computed and cached
- **Freshness**: Can be refreshed on-demand by authorized users
- **Consistency**: Always reflects the structure of the source query
- **Staleness tracking**: Metadata tracks when it was last updated

## How It Works

```
table_taxa (taxa DB)
    ↓
    [Materialized View Query]
    ↓
table_idtax (main DB) ← Can be refreshed by data managers
```

The materialized view is refreshed using the `REFRESH MATERIALIZED VIEW CONCURRENTLY` command, which:
- Does not block reads during refresh
- Requires a unique index (automatically created during migration)
- Updates data from the source query

---

# Key Functions

## 1. Checking Staleness: `check_table_idtax_staleness()`

Check if `table_idtax` needs to be refreshed based on how long ago it was last updated.

### Usage

```{r}
library(CafriplotsR)

# Connect to database
con <- call.mydb()

# Check staleness (warns if > 90 days old)
staleness <- check_table_idtax_staleness(con)

# Check staleness with custom threshold
staleness <- check_table_idtax_staleness(con, warn_days = 30)

# Silent check (no messages)
staleness <- check_table_idtax_staleness(con, silent = TRUE)
```

### Return Value

Returns a list with:

- `is_stale`: Logical, TRUE if older than threshold
- `days_old`: Numeric, age in days
- `last_updated`: POSIXct timestamp of last refresh
- `message`: Character, human-readable status message

### Example Output

```{r}
# Example 1: Fresh data
# $is_stale
# [1] FALSE
#
# $days_old
# [1] 15
#
# $last_updated
# [1] "2026-12-28 10:23:45 UTC"
#
# $message
# [1] "table_idtax was refreshed 15 days ago (2026-12-28 10:23:45)"

# Example 2: Stale data (with warning)
# ✖ Warning: table_idtax is stale (120 days old). Consider refreshing.
# $is_stale
# [1] TRUE
#
# $days_old
# [1] 120
```

### Practical Use

```{r}
# Check before running taxonomic queries
staleness <- check_table_idtax_staleness(con, warn_days = 60)

if (staleness$is_stale) {
  message("Consider refreshing table_idtax before proceeding.")
  # Optionally refresh automatically
  # update_taxa_link_table(con)
}
```

---

## 2. Refreshing the View: `update_taxa_link_table()`

Refresh `table_idtax` to synchronize with the latest data from `table_taxa`.

### Usage

```{r}
# Connect to both databases
con <- call.mydb()
con_taxa <- call.mydb.taxa()

# Refresh if stale (checks staleness first)
result <- update_taxa_link_table(con, con_taxa)

# Force refresh regardless of staleness
result <- update_taxa_link_table(con, con_taxa, force = TRUE)

# Custom staleness threshold (default 90 days)
result <- update_taxa_link_table(con, con_taxa, warn_days = 30)
```

### Behavior

1. **Checks staleness first** (unless `force = TRUE`)
   - If recently updated, skips refresh and returns success message
   - Threshold controlled by `warn_days` parameter

2. **Attempts materialized view refresh**
   - Calls PostgreSQL function `refresh_table_idtax()`
   - Uses `REFRESH MATERIALIZED VIEW CONCURRENTLY` (non-blocking)
   - Updates metadata table with timestamp and record count

3. **Fallback to legacy method** (if function not available)
   - Re-fetches data from `table_taxa` via R
   - Recreates the table (requires admin privileges)

### Return Value

Returns a list with:

- `success`: Logical, TRUE if refresh succeeded
- `method`: Character, "materialized_view", "legacy", or "skipped"
- `message`: Character, status message
- `record_count`: Integer, number of records after refresh
- `duration`: difftime, time taken for refresh

### Example Output

```{r}
# Example 1: Successful refresh
# ✔ Successfully refreshed table_idtax via materialized view
# ℹ 362,093 records refreshed in 2.3 seconds
#
# $success
# [1] TRUE
#
# $method
# [1] "materialized_view"
#
# $record_count
# [1] 362093
#
# $duration
# Time difference of 2.3 secs

# Example 2: Skipped (recently updated)
# ℹ table_idtax was refreshed 5 days ago. Skipping refresh.
# ℹ Use force = TRUE to refresh anyway.
#
# $success
# [1] TRUE
#
# $method
# [1] "skipped"
```

### Permissions

To refresh the materialized view, you need one of:

1. **SECURITY DEFINER function privilege**: The `refresh_table_idtax()` function is created with `SECURITY DEFINER`, allowing it to execute with the creator's (admin's) privileges
2. **data_manager_role**: If granted by the admin (requires `CREATEROLE` privilege)
3. **Direct REFRESH privilege**: On the materialized view

Most data managers should have access via the SECURITY DEFINER function.

---

## 3. Viewing Metadata: `get_table_idtax_metadata()`

Retrieve metadata about `table_idtax` refresh history.

### Usage

```{r}
con <- call.mydb()

# Get metadata
metadata <- get_table_idtax_metadata(con)
```

### Return Value

Returns a data.frame with:

- `table_name`: Character, always "table_idtax"
- `last_updated`: POSIXct, timestamp of last refresh
- `updated_by`: Character, username who performed refresh
- `record_count`: Integer, number of records after refresh
- `source_info`: Character, refresh method or migration note

### Example Output

```{r}
#   table_name       last_updated        updated_by record_count
# 1 table_idtax 2026-12-28 10:23:45           dauby       362093
#                                          source_info
# 1 Refreshed via refresh_table_idtax() function
```

### Practical Use

```{r}
# Check who last refreshed and when
metadata <- get_table_idtax_metadata(con)
cat(sprintf(
  "Last refresh: %s by %s (%s records)\n",
  metadata$last_updated,
  metadata$updated_by,
  format(metadata$record_count, big.mark = ",")
))
```

---

# Complete Workflow Example

## Scenario: Monthly Data Manager Check

```{r}
library(CafriplotsR)

# 1. Connect to databases
con <- call.mydb()
con_taxa <- call.mydb.taxa()

# 2. Check staleness
staleness <- check_table_idtax_staleness(con, warn_days = 30)

# 3. Refresh if needed
if (staleness$is_stale) {
  cat("Table is stale. Refreshing...\n")
  result <- update_taxa_link_table(con, con_taxa)

  if (result$success) {
    cat(sprintf("✓ Successfully refreshed %s records\n",
                format(result$record_count, big.mark = ",")))
  } else {
    cat("✗ Refresh failed. Contact admin.\n")
  }
} else {
  cat("Table is up to date. No refresh needed.\n")
}

# 4. View refresh history
metadata <- get_table_idtax_metadata(con)
print(metadata)

# 5. Clean up
cleanup_connections()
```

---

# Troubleshooting

## Error: "permission denied for materialized view table_idtax"

**Cause**: Your user account does not have permission to refresh the materialized view.

**Solution**: Ask the database administrator to:

1. Grant you the `data_manager_role`:
   ```sql
   GRANT data_manager_role TO your_username;
   ```

2. Or grant direct permission:
   ```sql
   GRANT SELECT, TRIGGER ON table_idtax TO your_username;
   ```

## Error: "function refresh_table_idtax() does not exist"

**Cause**: The migration to materialized view has not been completed.

**Solution**: Contact the database administrator to run:

```{r}
con <- call.mydb()
con_taxa <- call.mydb.taxa()

migrate_table_idtax_to_materialized_view(
  con = con,
  con_taxa = con_taxa,
  data_manager_users = c("dauby", "alex", "libalah")
)
```

## Warning: "table_idtax is stale"

**Cause**: The materialized view has not been refreshed recently.

**Solution**: Run `update_taxa_link_table()` to refresh:

```{r}
con <- call.mydb()
con_taxa <- call.mydb.taxa()
update_taxa_link_table(con, con_taxa)
```

## `update_taxa_link_table()` returns "skipped" but you want to force refresh

**Solution**: Use the `force = TRUE` parameter:

```{r}
update_taxa_link_table(con, con_taxa, force = TRUE)
```

---

# Background Information

## What Happens During Refresh?

1. **PostgreSQL locks the view** for writing (reads still allowed with `CONCURRENTLY`)
2. **Executes the source query** on `table_taxa` in the taxa database
3. **Replaces old data** with new query results
4. **Updates metadata table** with timestamp, record count, and username
5. **Releases lock** and returns success

## Performance Considerations

- **Refresh time**: Typically 2-5 seconds for ~360,000 records
- **No downtime**: Queries continue to work during refresh
- **Indexes maintained**: Unique index on `idtax` is automatically updated
- **Metadata overhead**: Minimal (<1ms)

## Automatic Refresh (Optional)

If the `pg_cron` extension is enabled by the administrator, `table_idtax` can be automatically refreshed on a schedule (e.g., monthly on the 1st at 2:00 AM):

```sql
-- Requires superuser privileges and pg_cron extension
SELECT cron.schedule(
  'refresh_table_idtax_monthly',
  '0 2 1 * *',
  'SELECT refresh_table_idtax();'
);
```

Check with your administrator if automatic refresh is configured.

---

# Summary

| Function | Purpose | Permissions Required |
|----------|---------|---------------------|
| `check_table_idtax_staleness()` | Check if refresh needed | Read access to metadata table |
| `update_taxa_link_table()` | Refresh the materialized view | REFRESH privilege or data_manager_role |
| `get_table_idtax_metadata()` | View refresh history | Read access to metadata table |

**Best Practice**: Check staleness monthly and refresh if older than 30-90 days.

**Key Advantages**:
- Non-admin users can refresh taxonomy data
- Staleness warnings prevent using outdated synonyms
- Metadata provides audit trail of refreshes
- Concurrent refresh allows queries during update

---

# Further Reading

- PostgreSQL Materialized Views: [https://www.postgresql.org/docs/current/rules-materializedviews.html](https://www.postgresql.org/docs/current/rules-materializedviews.html)
- Migration plan: `inst/docs/PLAN_MATERIALIZED_VIEW_MIGRATION.md`
- Quick reference: `inst/docs/QUICK_REFERENCE_table_idtax.md`
- Full SQL script: `inst/sql/migration_table_idtax_to_materialized_view.sql`
