---
title: "Using query_plots() to Access Plot Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using query_plots() to Access Plot Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,  # Set to FALSE to prevent execution during pkgdown build
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(tidyverse)
library(plotsdatabase)
```

## Introduction

The `query_plots()` function is the primary entry point for accessing forest plot data from the Central African plot database. This vignette demonstrates how to use the function's various features, including the new **output styles system** introduced in version 1.5.

## Database Connection

Before querying data, establish connections to both databases:


```{r connection-t, eval=T, echo=F, warning=FALSE, message=FALSE}
library(tidyverse)
devtools::load_all()
# Connect to the main database
mydb <- call.mydb(use_env_credentials = T)

# Connect to the taxa database
mydb_taxa <- call.mydb.taxa(use_env_credentials = T)

```

```{r connection}
# Connect to the main database
mydb <- call.mydb()

# Connect to the taxa database
mydb_taxa <- call.mydb.taxa()
```

The package will prompt for credentials interactively, or you can use environment variables via `use_env_credentials = TRUE`.

## Basic Usage

### Query Plot Metadata Only

To retrieve only plot-level information without individual tree data:

```{r basic-metadata, eval=TRUE, results='hide'}
# Query by plot name
plots <- query_plots(
  plot_name = "mbalmayo001",
  extract_individuals = FALSE
)

```

```{r, eval=TRUE, message=TRUE}
# View the structure
plots
```



### Query Plots with Individual Trees

To include individual tree/stem measurements:

```{r with-individuals, eval=TRUE, results='hide'}
plots_with_trees <- query_plots(
  plot_name = "mbalmayo001",
  extract_individuals = TRUE
)


```


```{r, eval=TRUE, message=TRUE}
# The result is a structured list
plots_with_trees
```

## Output Styles System (v1.5+)

Starting in version 1.5, `query_plots()` returns **structured lists** instead of flat data frames. The structure adapts automatically based on the inventory method.

### Available Output Styles

- **`minimal`**: Essential plot metadata only, no features or traits
- **`standard`**: Standard analysis output with common features (default for unknown methods)
- **`permanent_plot`**: Organized tables for permanent plot monitoring (single census)
- **`permanent_plot_multi_census`**: Multi-census data with _census_N columns
- **`transect`**: Simplified output for transect/walk surveys
- **`full`**: Complete export with all columns (preserves v1.4 behavior)

### Auto-Detection Based on Method

The output style is automatically detected from the plot's `method` field:

```{r auto-detection}
# For a permanent plot (method = "1 ha plot")
perm_plot <- query_plots(
  plot_name = "mbalmayo",
  extract_individuals = TRUE
)

# Automatically uses 'permanent_plot' style
# Returns: $metadata, $individuals, $censuses, $height_diameter
names(perm_plot)
```

### Manual Style Selection

Override automatic detection by specifying `output_style`:

```{r manual-style}
# Force minimal output
minimal_output <- query_plots(
  plot_name = "mbalmayo",
  extract_individuals = TRUE,
  output_style = "minimal"
)

# Force full output (v1.4 behavior - flat data frame)
full_output <- query_plots(
  plot_name = "mbalmayo",
  extract_individuals = TRUE,
  output_style = "full"
)
```

### Understanding the Structured Output

For permanent plots with `extract_individuals = TRUE`:

```{r structured-output}
result <- query_plots(
  plot_name = "mbalmayo001",
  extract_individuals = TRUE
)

# Metadata table: plot-level information
result$metadata
#   plot_id | plot_name | country | latitude | longitude | elevation | ...

# Individuals table: tree measurements
result$individuals
#   id_n | plot_name | tag | family | genus | species | dbh | height | ...

# Censuses table: census information
result$censuses
#   plot_name | census_number | census_date | team_leader | ...

# Height-diameter pairs: for allometric analysis
result$height_diameter
#   id_n | plot_name | tag | D | H | POM | census

# Spatial data (if show_all_coordinates = TRUE)
result$coordinates_sf
```

## Filtering Options

### By Location

```{r filter-location}
# By country
cameroon_plots <- query_plots(
  country = "Cameroon",
  extract_individuals = FALSE
)

# By locality
dja_plots <- query_plots(
  locality_name = "Dja",
  extract_individuals = FALSE
)

```

### By Method

```{r filter-method}
# Get all permanent plots
permanent_plots <- query_plots(
  method = "1 ha plot",
  extract_individuals = FALSE
)

# Get all transects
transects <- query_plots(
  method = "Long Transect",
  extract_individuals = FALSE
)
```

### By Plot ID

```{r filter-id}
# Query specific plot IDs
plots_by_id <- query_plots(
  id_plot = c(1, 2, 3),
  extract_individuals = TRUE
)
```

## Working with Individual Tree Data

### Basic Tree Extraction

```{r individuals-basic}
trees <- query_plots(
  plot_name = "mbalmayo001",
  extract_individuals = TRUE
)

# Access individual tree data
trees$individuals
```


## Handling Multiple Censuses

### Show Multiple Census Data

By default, only the most recent census is returned. To see all censuses:

```{r multiple-census}
multi_census <- query_plots(
  plot_name = "mbalmayo",
  extract_individuals = TRUE,
  show_multiple_census = TRUE
)

# Automatically uses 'permanent_plot_multi_census' style
# Census-specific columns: dbh_census_1, dbh_census_2, height_census_1, ...
head(multi_census$individuals)

# Censuses table shows all census dates
multi_census$censuses

# Height-diameter table has one row per measurement
multi_census$height_diameter
```

## Column Renaming

The output styles automatically rename database columns to user-friendly names:

```{r column-names}
result <- query_plots(
  plot_name = "mbalmayo001",
  extract_individuals = TRUE
)

# Database names → User-friendly names
# ddlat → latitude
# ddlon → longitude
# tax_fam → family
# tax_gen → genus
# tax_sp_level → species
# stem_diameter → dbh
# tree_height → height

names(result$metadata)
names(result$individuals)
```

## Spatial Data

### Plot Coordinates

By default, returns one coordinate per plot:

```{r default-coords}
plots <- query_plots(
  country = "Gabon",
  extract_individuals = FALSE
)

plots$metadata[, c("plot_name", "latitude", "longitude")]
```

### All Coordinates

Some plots have multiple coordinate points (corners, subplot centers):

```{r all-coords}
plots_all_coords <- query_plots(
  plot_name = "mbalmayo",
  extract_individuals = FALSE,
  show_all_coordinates = TRUE
)

# Returns sf object with all coordinate points
plots_all_coords$coordinates_sf
```

### Interactive Map

Visualize plot locations:

```{r map}
query_plots(
  country = "Cameroon",
  extract_individuals = FALSE,
  map = TRUE
)
```

## Advanced Options

### Keeping Database IDs

By default, internal IDs are removed. To keep them:

```{r keep-ids}
with_ids <- query_plots(
  plot_name = "mbalmayo001",
  extract_individuals = TRUE,
  remove_ids = FALSE, output_style = "full"
)

# Now includes id_liste_plots, id_census, etc.
names(with_ids$extract)
```

### Filtering by Individual ID

Extract specific individuals:

```{r filter-individual}
specific_trees <- query_plots(
  id_individual = c(1001, 1002, 1003),
  extract_individuals = TRUE
)
```

### Filtering by Taxon

Extract individuals of specific taxa:

```{r filter-taxon}
species_trees <- query_plots(
  plot_name = "mbalmayo001",
  id_tax = c(115),  # Taxon IDs
  extract_individuals = TRUE
)
```

## Custom Output Styles

The `keep_patterns` and `remove_patterns` options in style configurations allow customization (NOY YET IMPLEMENTED):

```{r custom-patterns}
# In your custom style configuration:
# keep_patterns = c("^feat_wood", "^trait_leaf")
# This keeps wood features and leaf traits while removing others
```

## Combining Options

Complex queries combining multiple filters:

```{r complex-query}
complex_result <- query_plots(
  country = "Cameroon",
  locality_name = "Dja",
  method = "1 ha plot",
  extract_individuals = TRUE,
  extract_traits = TRUE,
  extract_individual_features = TRUE,
  show_multiple_census = TRUE,
  show_all_coordinates = TRUE,
  remove_ids = FALSE,
  output_style = "permanent_plot_multi_census"
)

# Returns comprehensive structured output
names(complex_result)
```






## Performance Tips

1. **Start with metadata only**: Use `extract_individuals = FALSE` when exploring
2. **Use specific filters**: Reduce query time by filtering by plot_name or id_plot
3. **Limit traits**: Only use `extract_traits = TRUE` when needed
4. **Choose appropriate style**: Use `minimal` or `standard` for quick analyses

## Troubleshooting

### Empty Results

If your query returns no data:

```{r troubleshoot}
# Check connection status
print_connection_status()

# Run diagnostic
db_diagnostic()

# Verify plot access (row-level security)
# You can only access plots you have permission for
```

### Switching Back to v1.4 Behavior

If you need the old flat data frame output:

```{r legacy-behavior}
flat_output <- query_plots(
  plot_name = "mbalmayo001",
  extract_individuals = TRUE,
  output_style = "full"
)

# Returns a single data frame instead of structured list
```

