---
title: "Traits sampling summary in Mbalmayo, Bouamir and Loundoungou"
author: "Gilles Dauby"
date: "`r Sys.Date()`"
output:
  html_document:
  toc: true
toc_float: true
beamer_presentation: default
slidy_presentation: default
ioslides_presentation: default
pdf_document: default
---



```{r setup, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
devtools::load_all()
mydb <- call.mydb(use_env_credentials = T)
mydb_taxa <- call.mydb.taxa(use_env_credentials = T)
library(tidyverse)
library(gt)
library(sf)
library(ade4)
library(RecordLinkage)
```

```{r PackNeed, include=TRUE, eval=FALSE, echo=TRUE}
# to install the package 'cafriplotsR'
# make sure you are running recent R version

install.packages(c("tidyverse", "dbplyr", "devtools"))
devtools::install_github("umr-amap/cafriplotsR", upgrade = "never")
## if the installation above failed, it may be because your timeout is too low
## you can increase the timeout by using the following code
options(timeout = 1000)

library(tidyverse)
library(cafriplotsR) ## R package for exploring the database
library(sf)
library(gt)

mydb <- call.mydb()
mydb_taxa <- call.mydb.taxa()

```

Vérifier que la version du package est bien la [dernière en ligne](https://github.com/umr-amap/cafriplotsR/blob/master/DESCRIPTION).


```{r, echo=TRUE, include=TRUE, eval=FALSE, message=FALSE}
packageVersion("cafriplotsR")
```


# Extract data from the database

Concerning observatories, taxonomic and traits data are stored and managed in this database. Phenological characterization and spatial vectors linked to each tree (crown polygons) are managed trough the Shiny apps.

Data are stored in a postrgresql relational database and are available with several 'query' functions using an R package and a proper password and login name (ask Gilles).

Data is subject to evolve regularly : new data can be added and taxonomic identification can be updated.

Extract datasets for mbalmayo and bouamir observatories.

```{r extractData, echo=TRUE, include=TRUE, eval=TRUE, message=FALSE, cache=F, results='hide'}

tree_data <-
  query_plots(
    ### Extract from those localityes
    locality_name = c("Bouamir", "Mbalmayo", "Loundoungou"),
    ### Extract inventories of these specific methods
    method = c("1ha-IRD", "phenological_observatory"),
    ### Get inventories
    extract_individuals = TRUE,
    ### keeping keys
    remove_ids = FALSE, 
    ### output styl as full (many columns)
    output_style = "full"
  )

metadata <- 
  tree_data$metadata

tree_data <- 
  tree_data$extract

```



Query all features (including traits) linked to an individual in the inventories. A feature can be anything (including traits but not only).
```{r extractFeatMba, echo=TRUE, include=TRUE, eval=TRUE, cache=F, results='hide'}

all_traits <- 
  traits_list()

selected_traits <- 
  all_traits %>% 
  filter(grepl("lma_", trait) |
           trait %in% c(
             "twig_specific_gravity",
             "leaf_area",
             "SPAD_chlorophyll_measurement",
             "mesophyl_thickness",
             "whole_leaf_area",
             "leaf_structure_mass_area",
             "total_chlorophyll_content",
             "leaf_delta_carbon_13",
             "leaf_N_percentage",
             "leaf_C_percentage"
           ))

### export all features linked to an individual

all_traits_features <-
  query_individual_features(
    individual_ids = tree_data$id_n, 
    format = "long", 
    remove_issues = TRUE, 
    include_metadata = TRUE,
    trait_ids = selected_traits$id_trait)



```

# Explore and rearrange the dataset

Here we only want want features that are numeric, so we filter out everything that is not numeric and keep it in a dedicated table.
```{r featsOutputs, echo=TRUE, include=TRUE, eval=TRUE, message=FALSE}

## extracting traits/features that are numeric
numeric_traits <- 
  traits_list() %>% filter(valuetype == "numeric")

## Filtering out all traits/features to keep only numeric ones
all_features_num <- 
  all_traits_features %>% 
  filter(id_trait %in% numeric_traits$id_trait)


### count number of values for all numerical features

list_features_num <-
  all_features_num %>%
  group_by(trait, id_trait) %>%
  count()


```


# How traits are collected in the field and metada are associated

Each set of traits measured on a tree in an observatory is associated with a unique number __pheno_tree_nbr__ which is incremental. 

Thus, each __pheno_tree_nbr__ is linked to a tree, to trait measurements, to any observations/remarks on these measurements (for example the *id* of the leaf, because leaf traits are measured for at least 3 leaves), to any axillary measures used for the measurements (for example the area of the leaf for leaf mass per area measurements), to people involved in the measurements and to the date of measurement.
```{r ExtractPHTraits, echo=TRUE, include=TRUE, eval=TRUE}

### Extract all measurements with non empty pheno_tree_nbr (traits measured in Cameroon observatories)
all_traits_canopy_trees_cameroon <- 
  all_features_num %>%
  filter(!is.na(pheno_tree_nbr)) %>% ### non empty pheno_tree_nbr
  left_join( ### joining and retreiving tree dataset
    tree_data %>% select(id_n, tax_sp_level, tax_gen, plot_name, tag),
    by = c("id_data_individuals" = "id_n")
  ) %>% 
  filter(!grepl("lound", plot_name)) %>% 
  mutate(colnbr = pheno_tree_nbr) %>% 
  mutate(voucher_type = "phenotree")

### Extract all measurements with non empty pheno_tree_nbr (traits measured in Congo obsevatory)
all_traits_canopy_trees_congo <- 
  all_features_num %>%
  filter(!is.na(pheno_tree_nbr)) %>% ### non empty pheno_tree_nbr
  left_join( ### joining and retreiving tree dataset
    tree_data %>% select(id_n, tax_sp_level, tax_gen, plot_name, tag),
    by = c("id_data_individuals" = "id_n")
  ) %>% 
  filter(grepl("lound", plot_name)) %>% 
  mutate(colnbr = pheno_tree_nbr) %>% 
  mutate(voucher_type = "phenocongo")

## combining both dataset
all_traits_canopy_trees <- 
  bind_rows(all_traits_canopy_trees_cameroon,
            all_traits_canopy_trees_congo)
### Here one row is one measurement


## Getting measurements done on trees from plots
all_traits_plot <-
  all_features_num %>%
  select(-id_specimen) %>% 
  left_join(
    ### joining and retrieving tree dataset
    tree_data %>% select(
      id_n,
      tax_sp_level,
      tax_gen,
      plot_name,
      tag,
      id_specimen,
      colnbr
    ),
    by = c("id_data_individuals" = "id_n")
  ) %>%
  filter(!grepl("_observatory", plot_name)) %>% ### filter out all traits measured on trees from phenological observatories
  mutate(voucher_type = "pird")

## Combining all traits from plots and observatories
all_traits <- 
  bind_rows(all_traits_canopy_trees, all_traits_plot)


```

## Arranging the data into columns
```{r DataWide, echo=TRUE, include=TRUE, message=FALSE, eval=TRUE}

## retrieving information associated to each leaf : maturity and cohort when available
leaves_data <- 
  all_traits %>% 
  group_by(colnbr, voucher_type, leaf_id) %>% 
  summarise(leaf_maturity = first(leaf_maturity, na_rm = T),
            cohort_id = first(cohort_id, na_rm = T)) %>% 
  filter(!is.na(leaf_id)) %>% 
  ungroup()

## re arrangin the data in columns
# each trait columns will start with a prefix 't_' that help selection
all_traits_wide <-
  all_traits %>%
  select(
    id_data_individuals,
    id_trait_measures,
    colnbr,
    voucher_type,
    leaf_id,
    tag,
    plot_name,
    tax_sp_level,
    tax_gen,
    trait,
    traitvalue,
    day,
    month,
    year
  ) %>%
  pivot_wider(names_from = trait, 
              values_from = traitvalue,
              names_prefix = "t_")

traits_leaf_level <-
  all_traits_wide %>%
  ## grouping by voucher, tag, leaf etc..
  group_by(voucher_type ,
           colnbr,
           leaf_id,
           tag,
           tax_sp_level,
           id_data_individuals,
           plot_name) %>%
  ## get the mean for all traits and keeping the information for measurement date and plot
  summarise(across(starts_with("t_"), ~ mean(., na.rm = T)), 
            plot_name = first(plot_name),
            day = first(day, na_rm = T),
            month = first(month, na_rm = T),
            year = first(year, na_rm = T)) %>%
  ## joining leaves data that inform on the leaves where traits were measured
  left_join(leaves_data) %>%
  ungroup()

traits_leaf_level_averaged <-
  traits_leaf_level %>%
  ### grouping by voucher, tag, cohort, etc..
  group_by(voucher_type,
           colnbr,
           tag,
           tax_sp_level,
           cohort_id,
           leaf_maturity,
           id_data_individuals) %>%
  ## get the mean for all traits and keeping the information for measurement date and plot
  summarise(across(starts_with("t_"), ~ mean(., na.rm = T)),
            day = first(day, na_rm = T),
            month = first(month, na_rm = T),
            year = first(year, na_rm = T), 
            plot_name = first(plot_name), 
            plot_name = first(plot_name))


```

# Species level


## Grouping by species
```{r arrangeSpLevel, echo=TRUE, include=TRUE, eval=TRUE}

colnbr_per_sp <-
  traits_leaf_level %>%
  filter(leaf_maturity != "immature" |
           is.na(leaf_maturity)) %>%  ## excluding immature leaf for species level
  group_by(voucher_type, colnbr, tag, id_data_individuals, plot_name) %>%
  summarise(
    tax_sp_level = first(tax_sp_level),
    across(starts_with("t_"), ~ mean(., na.rm = TRUE)),
    plot_name = first(plot_name)
  )

traits_taxa <-
  colnbr_per_sp %>%
  mutate(locality = case_when(
    grepl("mbalmayo", plot_name) ~ "mbalmayo",
    grepl("bouamir", plot_name) ~ "bouamir",
    grepl("loundoungou", plot_name) ~ "loundoungou"
  )) %>% 
  ## grouping by species
  group_by(tax_sp_level, locality) %>%
  summarise(across(
    starts_with("t_"),
    list(
      mean = ~ round(mean(.x, na.rm = TRUE), 2),
      sd   = ~ round(sd(.x, na.rm = TRUE), 2),
      n    = ~ sum(!is.na(.x))
    ),
    .names = "{.col}_{.fn}"
  )) %>%
  filter(!grepl("indet", tax_sp_level), !is.na(tax_sp_level))

writexl::write_xlsx(traits_taxa, "traits_taxa.xlsx")

```


```{r}

tree_ind_dataset <- 
  tree_data %>% 
  filter(grepl("bouamir0", plot_name)) %>% 
  select(tax_sp_level, plot_name)

joined_plot_traits <- 
  tree_ind_dataset %>% 
  left_join(traits_taxa %>%
              filter(locality == "bouamir") %>% 
  select(
    tax_sp_level,
    t_lma_lamina_mean,
    t_lma_mesophyll_mean,
    t_lma_whole_leaf_mean,
    t_mesophyl_thickness_mean,
    t_SPAD_chlorophyll_measurement_mean,
    t_leaf_area_mean,
    t_leaf_structure_mass_area_mean,
    t_twig_specific_gravity_mean,
    t_leaf_N_percentage_mean,
    t_leaf_C_percentage_mean
  )) %>% 
  arrange(tax_sp_level)

trait_cols <- grep("^t_", names(joined_plot_traits), value = TRUE)

prop_indiv_plot <- 
  joined_plot_traits %>%
  group_by(plot_name) %>%
  summarise(across(all_of(trait_cols),
                   ~ round(mean(!is.na(.x)) * 100, 1),
                   .names = "prop_indiv_{.col}"))

prop_species_plot <- 
  joined_plot_traits %>%
  group_by(plot_name, tax_sp_level) %>%
  summarise(across(all_of(trait_cols), ~ any(!is.na(.x))), .groups = "drop") %>%
  group_by(plot_name) %>%
  summarise(across(all_of(trait_cols),
                   ~ round(mean(.x) * 100, 1),
                   .names = "prop_species_{.col}"))

prop_indiv_plot_long <- prop_indiv_plot %>%
  pivot_longer(
    cols = starts_with("prop_indiv_"),
    names_to = "trait",
    values_to = "prop"
  ) %>%
  mutate(trait = sub("^prop_indiv_", "", trait))

ggplot(prop_indiv_plot_long, aes(
  x = reorder(plot_name, prop),
  y = prop # conversion en pourcentage
)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~ trait, scales = "free_y") +
  coord_flip() +
  scale_y_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, by = 20),
    labels = function(x) paste0(x, "%")
  ) +
  labs(
    x = "Plot",
    y = "Complétude (%)",
    title = "Complétude des traits par plot (individus)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.y = element_text(size = 10)
  )

prop_species_plot_long <- prop_species_plot %>%
  pivot_longer(
    cols = starts_with("prop_species_"),
    names_to = "trait",
    values_to = "prop"
  ) %>%
  mutate(trait = sub("^prop_species_", "", trait))

ggplot(prop_species_plot_long, aes(
  x = reorder(plot_name, prop),
  y = prop
)) +
  geom_col(fill = "darkseagreen") +
  facet_wrap(~ trait, scales = "free_y") +
  coord_flip() +
  scale_y_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, by = 20),
    labels = function(x) paste0(x, "%")
  ) +
  labs(
    x = "Plot",
    y = "Complétude (%)",
    title = "Complétude des traits par plot (espèces)"
  ) +
  theme_minimal(base_size = 13)


sum_stems_missing_traits <- joined_plot_traits %>% 
  filter(is.na(t_twig_specific_gravity_mean), !is.na(tax_sp_level)) %>% 
  group_by(plot_name, tax_sp_level) %>% 
  count() %>% 
  arrange(desc(n))

missing_species <- 
  sum_stems_missing_traits %>% 
  ungroup() %>% 
  group_by(tax_sp_level) %>% 
  summarise(n = sum(n), freq = n()) %>% 
  arrange(desc(n)) %>% 
  rename(nbe_stems = n,
         nbe_plots = freq)

writexl::write_xlsx(missing_species, "D:/Papiers-Bourses-colloques/sejours_etrangers_bourses_financements_missions/mission_bouamir_novembre_2025//missing_species_wd.xlsx")



```









